---
title: Quarto Fundamentals
author: Tucker Plumlee and Paul Liu
date: 2026-06-03T14:30
date-format: "dddd, MMMM D, YYYY; h:mmA"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: false
    theme: flatly
    syntax-highlighting: tango
    smooth-scroll: true
    link-external-newwindow: true
---

# What is Quarto and Why Use It?

You may have noticed that all of the course's tutorials are provided as HTML files

## What is Quarto?

Quarto is a **code-based publishing system** that allows you to integrate executable code and plain text into a single document. Think of it as a way to combine your analysis, visualizations, and narrative all in one place.

If you have used **R Markdown** before, Quarto will feel familiar. It is the next generation of R Markdown, with some syntax differences you may notice when comparing with the R Markdown files used for Day 1 walkthroughs. Most importantly, Quarto is not limited to R --- it also supports **Python**, **Julia**, and **Observable** (a JavaScript-based visualization environment)

Quarto can output documents in many formats, including:

-   HTML pages
-   PDFs
-   Word documents
-   Presentations, websites, books, and more

## Why Use Quarto?

Quarto embraces the concept of **literate programming** --- the idea that code, results, and narrative explanation should live together. This means your data preparation, visualizations, and analysis all exist in one document.

The practical benefit is **reproducibility**: if your data changes, or if you revise your analysis approach, you can simply re-render the document and everything updates automatically. No more copying and pasting tables or figures into a separate Word document.

## Let's Get Started

To begin, let's create a new Quarto document in RStudio:

1.  Go to **File → New File → Quarto Document...**
2.  Give the document a title and author name
3.  Select **HTML** as the output format
4.  Click **Create**

RStudio will generate a template `.qmd` file with some example content. Click the **Render** button (or use the keyboard shortcut `Ctrl+Shift+K` / `Cmd+Shift+K`) to see the HTML output.

::: {.callout-tip}
## Try It
Create a new Quarto document now. Render it to see the default output before we start making changes.
:::

You can also view the Quarto document that produced this tutorial by opening `07_Day-2_Quarto-Fundamentals.qmd`.

# Basic Writing/Editing

## Markdown Basics

Quarto documents use **Markdown** syntax for formatting text. Here are the essentials:

### Headings

Use `#` symbols to create headings. More `#` symbols mean smaller headings:

``` markdown
# Heading 1
## Heading 2
### Heading 3
#### Heading 4
```

### Bold and Italics

Text can be formatted using the following syntax:

``` markdown
*This text is italic*/_This text is italic_
**This text is bold**/__This text is bold__
***This text is bold and italic***
```

*This text is italic*
**This text is bold**
***This text is bold and italic***.

### Non-Executable Code

You can also easily format text in monospace font to indicate code (without execution): `` `mean(x)` `` renders as `mean(x)`.

Use triple backticks for code blocks (without execution):

```` markdown
```
summary(data)
```
````

### Bullets and Lists

Unordered/bulleted lists use `-`, `+`, or `*` with a space afterwards:

``` markdown
- First item
+ Second item
  - Sub-item
  - Another sub-item
* Third item
```

- First item
+ Second item
  - Sub-item
  - Another sub-item
* Third item

Ordered lists use numbers:

``` markdown
1. First step
2. Second step
3. Third step
```

1.  First step
2.  Second step
3.  Third step

### Tables

You can manually create tables using pipes and dashes:

``` markdown
| Column A | Column B | Column C |
|----------|----------|----------|
| Row 1    | Data     | Data     |
| Row 2    | Data     | Data     |
```

Which renders as:

| Column A | Column B | Column C |
|----------|----------|----------|
| Row 1    | Data     | Data     |
| Row 2    | Data     | Data     |

Typing all this out can be painful --- the following website allows you to create a table visually, then copy and paste the associated markdown: [https://www.tablesgenerator.com/markdown_tables](https://www.tablesgenerator.com/markdown_tables).

::: {.callout-important}
## Why Learn Markdown?
Markdown is worth learning beyond Quarto. It is the standard formatting syntax used across many platforms: GitHub, Jupyter notebooks, Slack, Reddit, and many others. It is also the primary way people now communicate with AI tools like ChatGPT and Claude, especially in the context of agentic coding where developers use markdown documents to control agent behavior. Knowing Markdown will make you more effective across many digital workflows.
:::

## Source vs. Visual Editor

RStudio offers two editing modes for Quarto documents:

**Source Editor** shows the raw Markdown syntax. You will see the `#` symbols, `**bold**` markers, and all other formatting directly. This is useful when you want precise control over formatting, when you are writing code-heavy documents, or when troubleshooting rendering issues.

**Visual Editor** provides a WYSIWG ("what you see is what you get") experience similar to Microsoft Word. Formatting is displayed as it will appear in the rendered document, and you can use toolbar buttons for common formatting. This is useful when writing narrative-heavy sections, working with tables, or when you are new to Markdown.

To switch between them, use the **Source** and **Visual** buttons at the top-left of the editor pane.

::: {.callout-tip}
## Try It
Switch between Source and Visual modes now. Try typing some Markdown in Source mode, then switch to Visual mode to see how it renders. Notice how the Visual Editor also provides a toolbar for inserting images, tables, and footnotes.
:::

## Adding and Executing Code Chunks

### What are Code Chunks?

Code chunks are blocks of executable code embedded in your document. They are the core feature that separates Quarto from a plain text document. A code chunk looks like this:

````markdown
```{{r}}
# Your R code goes here
summary(mtcars)
```
````

To **add a new code chunk**, you can:

-   Use the keyboard shortcut `Ctrl+Alt+I` (Windows) or `Cmd+Option+I` (Mac)
-   Click the green **+C** button in the editor toolbar
-   Manually type the triple backtick fencing with `r` in curly-braces (if you are adding a code chunk using Python you would use `{python}` instead)

### Executing Code Chunks

There are several ways to run code in your document:

1.  **One line at a time**: Place your cursor on a line and press `Ctrl+Enter` / `Cmd+Enter`
2.  **Run an entire chunk**: Click the green play button at the top-right corner of the chunk, or press `Ctrl+Shift+Enter` / `Cmd+Shift+Enter`
3.  **Run all prior chunks**: Click the downward-arrow play button at the top-right of a chunk to run all chunks above the current one (or use the shortcut `Alt+Ctrl+Shift+P`). This is also available from the **Run** menu.

```{r}
#| label: basic-example
# Create a simple vector and calculate its mean
values <- c(12, 15, 18, 22, 25, 30)
mean(values)
```

Notice that the output will show up both in your Console, and below the chunk (i.e., "inline"). 

::: {.callout-warning}
## Execution Order Matters!
Code chunks execute in the order they appear in the document when you render. If you run chunks out of order during interactive editing (for example, running a chunk that uses an object defined earlier in the document without first running the earlier chunk), you will get an error. When in doubt, use **Run All Chunks Above** to make sure everything is up to date. If your document will not render, execution order is one of the first things to check.
:::

## Inline Code Execution

You can also embed R code directly within your text using the syntax ```{r} code```. This is useful for reporting results within a sentence.

For example, writing:

``` markdown
The dataset contains `{r} nrow(mtcars)` observations and `{r} ncol(mtcars)` variables.
```

Produces: "The dataset contains `{r} nrow(mtcars)` observations and `{r} ncol(mtcars)` variables." (`mtcars` is an example dataset built into base R).

This is a powerful feature for creating dynamic reports --- by using inline code execution if the data changes, the numbers in your text update automatically.

::: {.callout-warning}
## Reminder: Execution Order
Inline code follows the same execution order rules as code chunks. Any object you reference in inline code must be created in a code chunk *above* that point in the document.
:::

# Formatting and Rendering

## Basics of Rendering



## YAML Options

### What is YAML?

The block at the very top of your Quarto document (between the `---` lines) is called the **YAML header**. YAML stands for "Yet Another Markup Language," and it uses a simple `key: value` pair structure to define settings for different parameters. The YAML header is where you set **document properties** (title, author, date) as well as **rendering options** (output format, themes, table of contents, etc.).

Here is an example YAML header for an HTML document:

``` yaml
---
title: "My Analysis Report"
author: "Your Name"
date: 6/3/2026
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
---
```
In addition to setting the title, author, and date of the document, this YAML header ensures that, when the document is rendered in HTML, a table of contents is included, all header sections are numbered, and everything is visually styled according to the [Bootswatch cosmo theme](https://bootswatch.com/cosmo/).

### Specifying Output Formats

You can set one or more output formats in the YAML header. To add a **PDF** format option alongside HTML:

``` yaml
format:
  html:
    toc: true
    theme: flatly
  pdf:
    toc: true
    documentclass: article
```

When multiple formats are specified, you can choose which one to render using the dropdown next to the Render button, or from the command line with `quarto render document.qmd --to pdf`.

::: {.callout-note}
## PDF Output Requirements
Rendering to PDF requires a [LaTeX](https://www.latex-project.org/) installation. If you do not have one, Quarto can install **TinyTeX** for you by running `quarto install tinytex` in your terminal.
:::

## Code Chunk Options

In Quarto code chunk options are set using the `#|` (hashpipe) syntax at the top of a chunk. These control how each chunk behaves during rendering. Below are some of the most common.

### Adding a Label

Labels help with navigation and cross-referencing. You can cross-reference a code chunk by using the `@label-name` syntaxt. If you prefix the label with `fig-` or `tbl-` Quarto will enable automatic numbering of figures and plots when cross-referencing (e.g., Table 1, Table 2, etc.):

```{{r}}
#| label: fig-example-plot
```

### Controlling Code Display with `echo`

The `echo` option controls whether the **source code** appears in the rendered document.

```{{r}}
#| label: echo-demo
#| echo: false

# This code runs, but will not appear in the rendered document
hidden_result <- 42 * 2
hidden_result
```

The above code chunk is included in an executable chunk below.

```{r}
#| label: echo-demo
#| echo: false

# This code runs, but will not appear in the rendered document
hidden_result <- 42 * 2
hidden_result
```

The result above (`r hidden_result`) was produced by code, but the code itself is hidden because `echo: false` was set.

### Controlling Code Execution with `eval`

The `eval` option controls whether the code **runs** at all:

```{r}
#| label: eval-demo
#| eval: false

# This code is displayed but does NOT execute
this_would_cause_an_error <- nonexistent_function()
```

The chunk above is displayed but was never executed, so no error occurs (which would ordinarily prevent rendering).

### Controlling Output Display with `include`

The `include` option controls whether **anything** from the chunk (code or results) appears in the rendered document. When `include: false`, the code still runs, but nothing is shown:

```{{r}}
#| label: include-demo
#| include: false

# This runs silently --- useful for setup code
library(ggplot2)
sample_data <- mtcars
sample_data
```
This code is included below, but no code or output are displayed.

```{r}
#| label: include-demo
#| include: false

# This runs silently --- useful for setup code
library(tidyverse)
sample_data <- mtcars
```

We can confirm that the code was run by using the inline snippet, `` `{r} nrow(sample_data)` ``: `{r} nrow(sample_data)`.

### Setting Global Options in the YAML Header

Instead of setting options chunk-by-chunk, you can set defaults for the entire document in the YAML header:

``` yaml
---
title: "My Report"
execute:
  echo: true
  warning: false
  message: false
---
```

This hides warnings and messages throughout the document while still displaying code. Individual chunks can override these defaults.

## Tables, Figures, and References

Let's create some example tables and figures using the built-in `mtcars` dataset.

### Table Display

By default, Quarto displays data frames based on the `df-print` option set in the YAML header. This document uses the default `df-print: kable`, which produces clean, formatted tables:

```{r}
#| label: tbl-mtcars-summary
#| tbl-cap: "Summary Statistics of MPG by Number of Cylinders"

# Create a summary table using base R
mtcars_summary <- mtcars |> 
  group_by(cyl) |> 
  summarize(
    across(
      c(mpg, hp, wt), 
      \(x) round(mean(x), 1)
    )
  ) |> 
  rename(
    "Cylinders" = cyl,
    "Avg MPG" = mpg,
    "Avg HP" = hp,
    "Avg Weight (1000 lbs)" = wt
  )

mtcars_summary
```

As shown in @tbl-mtcars-summary, vehicles with fewer cylinders tend to have higher fuel efficiency.

You can also format tables using dedicated packages for more control:

-   `knitr::kable()` --- simple, reliable table formatting
-   `kableExtra` --- extends kable with styling options
-   `flextable` --- flexible tables that work across output formats
-   `gt` --- a grammar of tables, great for publication-quality output

### Figure Display

You can control figure appearance using chunk options for dimensions, alignment, captions, and alt text:

```{r}
#| label: fig-mpg-scatter
#| fig-cap: "Relationship Between Vehicle Weight and Fuel Efficiency"
#| fig-alt: "Scatter plot showing a negative relationship between car weight and miles per gallon, with points colored by number of cylinders."
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: center

ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    x = "Weight (1,000 lbs)",
    y = "Miles per Gallon",
    color = "Cylinders"
  ) +
  theme_minimal(base_size = 13)
```

As shown in @fig-mpg-scatter, there is a clear negative relationship between vehicle weight and fuel efficiency. Note how Quarto automatically numbers the figure based on its order in the document.

#### Multiple Figures in One Chunk

When a code chunk produces multiple figures, you can control the layout using chunk settings like `fig-subcap` and `layout-ncol`:

```{r}
#| label: fig-mpg-panels
#| fig-cap: "Fuel Efficiency Distributions"
#| fig-subcap:
#|   - "Distribution of MPG"
#|   - "MPG by Cylinder Count"
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 4

# Panel (a): Histogram
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(x = "Miles per Gallon", y = "Count") +
  theme_minimal(base_size = 12)

# Panel (b): Boxplot
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "Number of Cylinders", y = "Miles per Gallon") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 12)
```

@fig-mpg-panels shows two views of the MPG distribution: @fig-mpg-panels-1 presents the overall distribution, while @fig-mpg-panels-2 breaks it down by cylinder count.

### Cross-References

For both figures and tables, you can create in-text references using `@label`. Quarto will automatically generate the correct numbering:

-   `@fig-mpg-scatter` renders as a clickable link like "@fig-mpg-scatter"
-   `@tbl-mtcars-summary` renders as "@tbl-mtcars-summary"

The labels must begin with the `fig-` or `tbl-` prefix, and the chunk must include a `fig-cap` or `tbl-cap` for cross-referencing to work.

## Footnotes

There are several ways to add footnotes in Quarto.

**Numbered footnotes** use a reference marker in the text and define the content elsewhere:

``` markdown
This is a statement that needs a footnote[^1].

[^1]: This is the footnote content.
```

This is a statement that needs a footnote[^1].

[^1]: This is the footnote content.

**Named footnotes** use a descriptive label instead of a number, which can be easier to manage in longer documents:

``` markdown
Quarto supports many output formats[^formats].

[^formats]: Including HTML, PDF, Word, revealjs slides, and more.
```

Quarto supports many output formats[^formats].

[^formats]: Including HTML, PDF, Word, revealjs slides, and more.

**Inline footnotes** let you write the content directly in the text without a separate definition:

``` markdown
This is convenient for short notes^[Like this one, which does not require a separate definition.].
```
This is convenient for short notes^[Like this one, which does not require a separate definition.].

# More Features and Formats

## Caching Results

For documents with long-running computations, you can **cache** results so that chunks only re-execute when their code changes.

Add caching to a single chunk:

```{{r}}
#| cache: true

# This long computation will only run once
result <- some_slow_function(data)
```

Or enable caching for the entire document in the YAML header:

``` yaml
execute:
  cache: true
```

When caching is enabled, Quarto creates a folder named `document-name_cache` containing the stored results. If the code in a cached chunk changes, that chunk (and any chunks that depend on it) will re-execute.

::: {.callout-caution}
## Caching Caveats
Caching can sometimes cause unexpected behavior if chunks depend on external data that has changed or on objects modified in other chunks. If you encounter strange results, try deleting the cache folder and re-rendering.
:::

## Callout Blocks

Callout blocks are a great way to draw attention to important information. Quarto supports five types:

::: {.callout-note}
## Note
Use for additional context or supplementary information.
:::

::: {.callout-warning}
## Warning
Use to flag potential problems or common mistakes.
:::

::: {.callout-important}
## Important
Use for critical information the reader must not miss.
:::

::: {.callout-tip}
## Tip
Use for helpful suggestions and best practices.
:::

::: {.callout-caution}
## Caution
Use to warn about actions that could cause issues.
:::

The syntax for creating a callout is:

``` markdown
::: {.callout-note}
## Optional Title
Content of the callout goes here.
:::
```

You can also add attributes. For example, a **collapsible** callout:

``` markdown
::: {.callout-tip collapse="true"}
## Click to Expand
This content is hidden by default.
:::
```

::: {.callout-tip collapse="true"}
## Click to Expand
This content is hidden by default. It appears when the reader clicks the callout header.
:::

You can also change the appearance to a simpler style:

``` markdown
::: {.callout-note appearance="simple"}
A simpler, more understated callout style.
:::
```

## Bibliography and Citation Integration

Quarto has built-in support for citations and bibliographies, making it an excellent tool for academic and research writing.

**Step 1**: Create or export a bibliography file (e.g., `references.bib`) from a reference manager like **Zotero**, **Mendeley**, or **EndNote**. A `.bib` file entry looks like this:

``` bibtex
@article{wickham2014,
  title={Tidy Data},
  author={Wickham, Hadley},
  journal={Journal of Statistical Software},
  volume={59},
  number={10},
  year={2014}
}
```

**Step 2**: Reference the bibliography file in your YAML header, and optionally specify a citation style:

``` yaml
---
title: "My Research Paper"
bibliography: references.bib
csl: apa.csl
---
```

Citation style files (`.csl`) for virtually any journal or style guide (APA, Chicago, MLA, etc.) can be downloaded from the [Zotero Style Repository](https://www.zotero.org/styles).

**Step 3**: Cite sources in your text using the `[@key]` syntax:

``` markdown
Tidy data principles have transformed data analysis workflows [@wickham2014].
```

Quarto will automatically generate a formatted **References** section at the end of your document.

## Dashboards, Websites, Books, and More!

Quarto can do much more than single-page reports. The same core skills you have learned today apply to a wide range of output types:

-   **Presentations**: Create slides using `format: revealjs` for HTML presentations or `format: beamer` for PDF slides
-   **Websites**: Build multi-page websites with navigation, hosted on GitHub Pages or other platforms
-   **Books**: Write multi-chapter books with cross-references, indexing, and multiple output formats
-   **Dashboards**: Create interactive dashboards using `format: dashboard` with layouts and value boxes

These all use the same Markdown syntax, code chunks, and YAML configuration you have practiced today.

::: {.callout-tip}
## Learn More
Visit the [Quarto documentation](https://quarto.org/docs/guide/) for guides and examples of all these formats.
:::

