---
title: "R Fundamentals"
author:
  - "Tucker Plumlee and Paul Liu"
  - "R Training for LMI Shops: Beginning to Intermediate"
date: "June 2, 2026: 9:30am - 10:30am"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This first session covers the basics you need to feel at home in R: what the windows in front of you are, how to run code, how to store things, how to add the tools (packages) we'll use, and how to load a data file. By the end you'll have read in the Oregon employment dataset we use for the rest of the day.

No prior programming experience is assumed. If something doesn't click right away, that's normal — you'll see all of it again in the hands-on sessions.

---

# 1. R and RStudio

**R** is the programming language that does the work. **RStudio** is the program you actually look at — it's an *IDE* (integrated development environment), which is just software that makes writing and running R easier: a place to write code, run it, see results, view plots, and read help.

When you open RStudio you'll see (up to) four **panes**:

- **Source** (top-left) — where you write and save scripts. If no file is open, this pane is hidden and the Console fills the space.
- **Console** (bottom-left) — where code actually runs and results appear.
- **Environment** (top-right) — a list of the objects (data, values) you've created.
- **Viewer/Output** (bottom-right) — plots, files, installed packages, and help pages.

You can rearrange and recolor these under **Tools > Global Options** if you like.

### Exercises

1. Open RStudio and find all four panes.
2. Open a new script with **File > New File > R Script**. Watch the Source pane appear.

---

# 2. The Console: Running Code

The Console runs code immediately. Type something after the `>` prompt and press Enter. R works as a calculator out of the box:

```{r}
2 + 2
100 / 4
sqrt(81)
```

A function is a command that does something; you give it inputs in parentheses:

```{r}
sum(10, 20, 30)
round(3.14159, 1)
```

The Console is great for quick, throwaway calculations. For anything you want to keep, use a script (next section).

### Exercises

1. Use the Console to compute `45 * 12`.
2. Try `round(2.71828, 2)`.

---

# 3. Scripts: Saving Your Work

A **script** is a plain text file (ending in `.R`) where you write code you want to keep. You type your code in the Source pane, then run it.

To run code from a script:

- **One line:** put your cursor on the line and press **Ctrl+Enter** (Cmd+Enter on Mac).
- **The whole script:** click **Source** at the top of the pane.

Anything after a `#` is a **comment** — R ignores it. Comments explain what your code does (to others, and to future you):

```{r}
# Calculate the average of three wages
mean(c(45000, 52000, 38000))
```

Save your script with **Ctrl+S**. Now your work is reproducible — you (or a colleague) can reopen it and run it again.

### Exercises

1. Open a new script, type the comment and `mean()` line above, and run it with Ctrl+Enter.
2. Save the script as `practice.R`.

---

# 4. Objects and Assignment

You store a value in an **object** using the assignment arrow `<-` (a less-than sign and a dash). RStudio has a shortcut: **Alt+-** (Alt and minus).

```{r}
wage <- 52000
wage
```

Once stored, you can use the object in later code:

```{r}
annual <- wage * 1
monthly <- wage / 12
monthly
```

Every object you create shows up in the **Environment** pane (top-right). Object names should be short and meaningful — use letters, numbers, and underscores, and don't start with a number.

### Exercises

1. Create an object `employees <- 250`.
2. Create `avg_wage <- 48000`, then compute total payroll as `employees * avg_wage`.

---

# 5. Vectors and Data Frames

A **vector** is a set of values of the same type, made with `c()` (for "combine"):

```{r}
wages <- c(45000, 52000, 38000, 61000)
wages
mean(wages)
max(wages)
```

A **data frame** is a table — rows and columns, like a spreadsheet. Each column is a vector. This is the shape almost all of our data takes. We won't build one by hand; we'll load one from a file in a moment, and every column in it (employment, wages, etc.) is just a vector.

### Exercises

1. Make a vector `emp <- c(10, 25, 5, 40)` and find its `sum()`.
2. Find the `mean()` of `emp`.

---

# 6. Packages

R comes with many built-in functions, but most of what we do uses **packages** — add-on collections of functions. Two steps, and people often confuse them:

- **Install once per computer** with `install.packages()`. (You've likely done this before the workshop.)
- **Load every session** with `library()`, at the top of your script.

```{r, eval=FALSE}
# Run these once, if you haven't already:
install.packages("tidyverse")
```

```{r}
# Run this at the start of every session:
library(tidyverse)   # data wrangling and plotting tools
```

The **tidyverse** is actually a bundle of packages (including `readr` for reading data and `dplyr` for wrangling). Loading it loads them all.

### Exercises

1. Load `tidyverse` and `here` with `library()`.
2. In your own words, what's the difference between `install.packages()` and `library()`?

---

# 7. Where Your Files Live: the Working Directory

Think of R like a chef in a kitchen. Before the chef can grab ingredients (like data files) or put down a finished dish (like a saved plot), they need to know which kitchen they are standing in.

In R, this "kitchen" is called your working directory—the specific folder on your computer where R automatically looks for files to read and where it will save files you create.

To find out exactly which folder R is currently looking at, you use the get working directory function:

```{r}
getwd()

```

When you run this, R will print out a "file path" (e.g., "C:/Users/YourName/Documents"). This is your current home base.

If your data files are in a different folder, you have to tell R to move to that folder using the set working directory function. You paste the file path inside the parentheses as a piece of text (in quotation marks):

```{r, eval=FALSE}
setwd("C:/Users/YourName/Documents/MyRFolder")

```

**Watch out for slashes!** Windows uses backslashes () for file paths, but R requires forward slashes (/) or double backslashes (\). If you copy and paste a path on Windows, you will need to change \ to /, or R will give you an error.

### Exercises

1. Run `getwd()` in your console. Write down or look closely at the folder path it returns.
2. Try using `setwd()` to change your working directory to that new folder. Run getwd() again to confirm that R successfully moved to the new location.

---

# 8. Reading In Data

```{r, echo=FALSE}
setwd("C:/Users/Paul Liu/Desktop/R Course (June 2-3)")
```


After setting your working directory using `setwd()`, we will use `read_csv()` (from the tidyverse) to load our first dataset into R. Let's load the Oregon QCEW employment data we'll use for the rest of the day:

```{r}
oregon_qcew <- read_csv("oregon_counties_qcew.csv")
```

Always look at what you just loaded:

```{r}
summary(oregon_qcew)   # summary stats of the variables in the data
glimpse(oregon_qcew)   # columns, types, first values
head(oregon_qcew)      # first six rows
```

You now have a data frame called `oregon_qcew` in your Environment. This is exactly where the next session — Data Wrangling — picks up.

### Exercises

1. Read in `oregon_counties_qcew.csv` as shown.
2. Run `glimpse()` on it. How many rows and columns does it have?

---

# 9. Good Habits

A few practices that save a lot of pain:

- **Write code in scripts, not just the Console.** The Console is for quick tests; the script is your saved, repeatable work.
- **Comment as you go** with `#`. Future-you will be grateful.
- **Start clean.** Restart R now and then (**Session > Restart R**) and re-run your script from the top. A script that runs top-to-bottom in a fresh session is a reliable script. One practice is to put **rm(list = ls())** at the top of your script. This clears all files in your global environment and allows your script to run with a blank slate.
- **Get help.** Put a `?` before any function name to open its help page:

```{r, eval=FALSE}
?mean
?read_csv
```

- **When stuck, read the error.** R's messages look intimidating but usually name the problem. Copying an error into a search engine almost always helps.

### Exercises

1. Open the help page for `read_csv` with `?read_csv`.
2. Restart R (**Session > Restart R**) and re-run your script from the top.

---

# Wrap-Up

You can now:

- recognize the **Source**, **Console**, **Environment**, and **Viewer** panes
- run code in the **Console** and save it in a **script**
- store values in **objects** with `<-`
- work with **vectors** and **data frames**
- **install** and **load** packages
- find files with an **RStudio Project** and **`here()`**
- **read in a CSV** and look at it

That last point — `oregon_qcew` loaded and ready — is the starting line for the Data Wrangling session next.

For reference and continued practice: [R for Data Science](https://r4ds.hadley.nz/) and the [Posit cheatsheets](https://posit.co/resources/cheatsheets/).
