---
title: "Data Wrangling with the Tidyverse"
author:
  - "Tucker Plumlee and Paul Liu"
  - "R Training for LMI Shops: Beginning to Intermediate"
date: "June 2, 2026: 10:45am - 12:00pm"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This section covers the core verbs of the **tidyverse** for working with data. We'll use a prepared extract of Oregon QCEW (Quarterly Census of Employment and Wages) data: private-sector employment and wages by county and industry for 2025.

Almost everything here comes from one package, **dplyr** (with a little **tidyr** for reshaping at the end). Loading `tidyverse` brings them both, plus the pipe.

# Setup

```{r, echo=FALSE}
rm(list=ls())
setwd("C:/Users/Paul Liu/Desktop/R Course (June 2-3)")
```


```{r}
library(tidyverse)

# The QCEW extract we'll work with. It has industry codes and county fips but not names —
# we'll add names later using a crosswalk, which is a nice excuse to learn joins.
oregon_qcew <- read_csv("oregon_counties_qcew.csv")

# Two small lookup tables for the joins section:
naics_crosswalk <- read_csv("NAICS_crosswalk.csv")

oregon_fips <- read_csv("oregon_fips.csv") %>%
  rename(county_name = `county name`)     # fips code -> county name

```

---

# 1. Looking at Your Data

Always look at a data frame before you start changing it. Three quick ways:

- `dim()` — how many rows and columns
- `names()` - column names in the dataset
- `glimpse()` — column names, types, and the first few values
- `summary()` - summary statistics of each column
- `View()` (capital V) — open a spreadsheet-style viewer in RStudio

```{r}
dim(oregon_qcew)
names(oregon_qcew)
summary(oregon_qcew)
glimpse(oregon_qcew)
# View(oregon_qcew)
```

Each row is one industry in one county:

- `county` — county name
- `naics` — 6-digit industry code (no name yet — we'll add that with a join)
- `establishments`, `avg_employment` — counts
- `avg_wkly_wage`, `avg_annual_wage` — dollars

### Exercises

1. Use `dim()` to confirm the number of rows and columns.
2. How many unique counties are in the data? (Hint: `unique()` then `length()`.)
3. Run `head(oregon_qcew)`. How is that different from `glimpse()`?

---

# 2. The Pipe Operator

The pipe `%>%` takes the thing on its left and feeds it into the function on its right. These two lines do exactly the same thing:

```{r}
head(oregon_qcew)
oregon_qcew %>% head()
```

The pipe is most useful when you do several things in a row. Read it top to bottom, like a recipe:

```{r}
oregon_qcew %>%
  filter(county == "Multnomah County, Oregon") %>%
  head()
```

### Exercises

1. Rewrite `mean(c(1, 2, 3, 4, 5))` using a pipe.
2. Pipe `oregon_qcew` into `nrow()`.

---

# 3. Choosing Columns with `select()`

`select()` keeps only the columns you name.

```{r}
oregon_qcew %>%
  select(county, naics, avg_employment) %>%
  head()
```

You can drop a column instead by putting a minus sign in front of it:

```{r}
oregon_qcew %>%
  select(-avg_annual_wage) %>%
  head()
```

### Exercises

1. Select just `county`, `naics`, and `avg_annual_wage`.
2. Select everything except `establishments`.

---

# 4. Choosing Rows with `filter()`

`filter()` keeps only the rows that meet a condition.

```{r}
oregon_qcew %>%
  filter(county == "Multnomah County, Oregon")
```

Use `>`, `<`, `>=`, `<=` for numbers:

```{r}
oregon_qcew %>%
  filter(avg_employment > 1000)
```

List several conditions separated by commas — a row has to meet *all* of them:

```{r}
oregon_qcew %>%
  filter(county == "Multnomah County, Oregon", avg_employment > 1000)
```

To keep rows matching any value in a list, use `%in%`:

```{r}
oregon_qcew %>%
  filter(county %in% c("Baker County, Oregon", "Benton County, Oregon"))
```

### Exercises

1. Filter to rows where `avg_annual_wage` is greater than 60000.
2. Filter to industries with more than 500 establishments.
3. Filter to two counties of your choice using `%in%`.

---

# 5. Creating Columns with `mutate()`

`mutate()` adds a new column calculated from existing ones.

```{r}
oregon_qcew %>%
  mutate(wage_thousands = avg_annual_wage / 1000) %>%
  select(county, naics, avg_annual_wage, wage_thousands) %>%
  head()
```

You can make several columns in one `mutate()`:

```{r}
oregon_qcew %>%
  mutate(wage_thousands = avg_annual_wage / 1000,
         emp_per_estab  = avg_employment / establishments) %>%
  head()
```

To create a column based on conditions, use `case_when()`. Each line is a condition on the left of the `~` and the value to assign on the right.

```{r}
oregon_qcew %>%
  mutate(size = case_when(
    avg_employment <  50                        ~ "Small",
    avg_employment >= 50 & avg_employment < 500 ~ "Medium",
    avg_employment >= 500                       ~ "Large"
  )) %>%
  select(county, naics, avg_employment, size) %>%
  head()
```

### Exercises

1. Create a column `total_payroll` equal to `avg_employment * avg_annual_wage`.
2. Create a column `wage_thousands` equal to `avg_annual_wage / 1000`.
3. Use `case_when()` to make a `wage_level` column: "Low", "Medium", or "High", using wage cutoffs you choose.

---

# 6. Renaming Columns with `rename()`

The pattern is `rename(new_name = old_name)`:

```{r}
oregon_qcew %>%
  rename(employment = avg_employment,
         annual_wage = avg_annual_wage) %>%
  head()
```

### Exercises

1. Rename `avg_wkly_wage` to `weekly_wage`.
2. Rename both `establishments` and `avg_employment` to names you like, in one `rename()` call.

---


# 7. Grouping and Summarizing

`group_by()` splits the data into groups, and `summarize()` collapses each group down to a single summary row. Together they answer questions like "what's the total for each county?"

```{r}
oregon_qcew %>%
  group_by(county) %>%
  summarize(total_emp = sum(avg_employment))
```

A few useful pieces:

- `sum()`, `mean()`, `n()` (counts rows in the group)
- `arrange()` sorts the result; wrap a column in `desc()` for high-to-low

```{r}
oregon_qcew %>%
  group_by(county) %>%
  summarize(total_emp    = sum(avg_employment),
            n_industries = n()) %>%
  arrange(desc(total_emp))
```

### Exercises

1. Find the average weekly wage for each county.
2. Count how many industries are in each county.
3. Find the five counties with the most total employment. (Hint: add `head(5)` at the end.)

---



# 8. Joining Tables

Often the information you need is split across two tables. A **join** matches them up on a shared column (a "key").

Our QCEW data has industry codes but no names. The crosswalk has both:

```{r}
naics_crosswalk %>% head()
```

`left_join()` keeps every row of the first table and adds matching columns from the second:

```{r}
oregon_qcew <- oregon_qcew %>%
  left_join(naics_crosswalk, by = "naics")

oregon_qcew %>%
  select(county, naics, industry_name, avg_employment) %>%
  head()
```

Same idea to add fips codes to the  county names from the FIPS lookup file, the data frame **oregon_fips**:

```{r}
oregon_qcew <- oregon_qcew %>%
  left_join(oregon_fips, by = c("county" = "county_name"))

oregon_qcew %>% select(fips, county, avg_employment) %>% head()
```

Now that we have the industry names, we can answer friendlier questions:

```{r}
oregon_qcew %>%
  group_by(industry_name) %>%
  summarize(total_emp = sum(avg_employment)) %>%
  arrange(desc(total_emp)) %>%
  head(10)
```

There are a few kinds of joins; `left_join()` is the one you'll reach for most:

- `left_join(x, y)` — keep all rows of `x`, add matches from `y`
- `inner_join(x, y)` — keep only rows that appear in both
- `full_join(x, y)` — keep everything from both

### Exercises

1. After the joins, which industry has the highest `avg_annual_wage`? (Hint: `arrange(desc(avg_annual_wage))`.)
2. Which county has the most total employment, using `county_name`?

---

# 9. Reshaping with Pivots

Sometimes data needs to change shape. The two tools are `pivot_longer()` (wide to long) and `pivot_wider()` (long to wide).

Our two wage columns hold the same kind of value (a wage) in two units. We can stack them into one column with `pivot_longer()`:

```{r}
wage_long <- oregon_qcew %>%
  select(county, industry_name, avg_wkly_wage, avg_annual_wage) %>%
  pivot_longer(cols      = c(avg_wkly_wage, avg_annual_wage),
               names_to  = "wage_type",
               values_to = "wage")

wage_long %>% head()
```

- `cols` — which columns to stack
- `names_to` — name of the new column holding the old column names
- `values_to` — name of the new column holding the values

`pivot_wider()` does the reverse, spreading one column back out into several:

```{r}
wage_long %>%
  pivot_wider(names_from  = wage_type,
              values_from = wage) %>%
  head()
```

### Exercises

1. In `wage_long`, how many rows are there compared to the original? Why?
2. Use `pivot_wider()` on `wage_long` to get back to one row per industry.

---

# 10. Putting It All Together

Here is a full, simple pipeline from the raw extract to a county summary — load, join, group, summarize. This is what a tidy script looks like in practice: a few steps, read top to bottom.

```{r}
county_summary <- oregon_qcew %>%
  group_by(county) %>%
  summarize(total_emp = sum(avg_employment),
            avg_wage  = round(mean(avg_annual_wage))) %>%
  arrange(desc(total_emp))

county_summary
```

This is the kind of table we'll turn into charts this afternoon in the ggplot2 session.

---

# Wrap-Up

You now know the core tidyverse verbs:

- **`select()`** — choose columns
- **`filter()`** — choose rows
- **`mutate()`** — add new columns (and `case_when()` for conditions)
- **`rename()`** — rename columns
- **`group_by()`** + **`summarize()`** — totals and averages by group
- **`left_join()`** — combine two tables on a shared key
- **`pivot_longer()`** / **`pivot_wider()`** — change the shape of data
- **`%>%`** — chain steps together

For more practice, the free online book [R for Data Science](https://r4ds.hadley.nz/) covers all of this, and the [dplyr cheat sheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf) is a handy one-page reference.

This afternoon: a hands-on wrangling exercise, then making charts with ggplot2.
