---
title: "Visualizing Data in R"
author:
  - "Tucker Plumlee and Paul Liu"
  - "R Training for LMI Shops: Beginning to Intermediate"
date: "June 2, 2026: 3:15am - 4:45pm"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

So far today, we've loaded data, cleaned it, reshaped it, and summarized it. But a summary table only gets you so far in a briefing or a slide deck. Now, we are going to make that data visual.

We use a package called ggplot2 (which loaded automatically when we ran library(tidyverse) this morning). It operates on a concept called the "Grammar of Graphics", which is the idea that you can build any chart by adding layers of data, aesthetics, and geometry on top of each other.

# 1. The Grammar of Graphics: Layer by Layer

Let's start with the county_tiers summary table we built in the applied exercise.

```{r}
# Load the raw data and the crosswalks
library(tidyverse)
oregon_qcew <- read_csv("oregon_counties_qcew.csv")
naics_crosswalk <- read_csv("NAICS_crosswalk.csv")
oregon_fips <- read_csv("oregon_fips.csv") %>% rename(county_name = `county name`)

# Join them together into one master table called 'or_data'
or_data <- oregon_qcew %>%
  left_join(naics_crosswalk, by = "naics") %>%
  left_join(oregon_fips, by = c("county" = "county_name"))

county_tiers <- or_data %>%
  filter(county != "Unknown Or Undefined") %>%
  group_by(county) %>%
  summarize(total_emp = sum(avg_employment)) %>%
  mutate(tier = case_when(
    total_emp < 10000 ~ "Under 10k",
    total_emp >= 10000 & total_emp <= 50000 ~ "10k-50k",
    total_emp > 50000 ~ "Over 50k"
  ))

county_tiers_count <- county_tiers %>%
  group_by(tier) %>%
  summarize(n_counties = n())

```


If you remember, the pipe %>% takes data and pushes it forward. In ggplot2, we use the + sign to add new layers to our canvas.

First, we set up the blank canvas and tell it what data to use:
```{r}
# 1. The "Canvas"
ggplot(data = county_tiers_count)

```


Next, we add aesthetics (aes()). This maps columns in our data to visual properties on the chart (like the x-axis, y-axis, or color)

```{r}
# 2. The Axes
ggplot(data = county_tiers_count, aes(x = tier, y = n_counties))

```


Finally, we add a geom (geometry) to tell R what type of visualization we want. Since we want a bar chart, we use geom_col().

```{r}
# 3. The Content
ggplot(data = county_tiers_count, aes(x = tier, y = n_counties)) +
  geom_col()

```

### Exercises

1. Change geom_col() to geom_point(). What happens?
2. Inside geom_col(), add the argument fill = "darkblue".


# 2. Visualizing Industries via a Horizontal Bar Chart

Let's look at the economic profile of Clatsop County.

First, let's use our wrangling skills to grab their top 10 industries by *total payroll*. We'll define this **as avg_employment x avg_wkly_wage**

```{r}
clatsop_top10_payroll <- or_data %>%
  filter(county == "Clatsop County") %>%
  mutate(payroll = avg_employment * avg_wkly_wage) %>% 
  group_by(industry_name) %>%
  summarize(total_payroll = sum(payroll, na.rm = TRUE)) %>%
  arrange(desc(total_payroll)) %>%
  head(10)

clatsop_top10_payroll

```


If we plot this normally, the long industry names will overlap on the x-axis and become completely unreadable:
```{r}
# Don't do this! The text will be an unreadable mess.
ggplot(clatsop_top10_payroll, aes(x = industry_name, y = total_payroll)) +
  geom_col()

```

Let's try the horizontal bar chart. We map the long names to the y axis, and use a  function called **reorder()** to sort the bars from largest to smallest.

```{r}
ggplot(clatsop_top10_payroll, 
       aes(x = total_payroll, y = reorder(industry_name, total_payroll))) +
  geom_col()

```


# 3. Formatting Visualizations

A grey background and raw variable names are fine for quick analysis, but you wouldn't drop this into a brief. Let's polish this chart for a final presentation.

We are going to add four things:

  - **theme_minimal()**: Cleans up the background.
  - **scale_x_continuous()**: Formats our raw numbers in dollars and with commas. (Requires the scales package).
  - **labs()**: Adds titles, subtitles, and cleans up the axis labels.
  
  
```{r}
# Load the scales package for number formatting
library(scales)

polished_plot <- ggplot(clatsop_top10_payroll, 
                        aes(x = total_payroll, y = reorder(industry_name, total_payroll))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = label_currency()) +
  theme_minimal() +
  labs(
    title = "Top 10 Industries by Total Estimated Payroll",
    subtitle = "Clatsop County, Oregon (2025)",
    x = "Total Estimated Payroll",
    y = NULL,
    caption = "Source: BLS QCEW; Q1-Q3 2025"
  )

polished_plot

```
  
The industry names on the y-axis is taking up too much of the page. We will use a function called **scale_y_discrete()** to automatically wrap the text on the y-axis after a certain number of characters.

```{r}
polished_plot <- ggplot(clatsop_top10_payroll, 
                        aes(x = total_payroll, y = reorder(industry_name, total_payroll))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = label_currency()) +
  scale_y_discrete(labels = label_wrap(30)) + # this is the new layer that wraps the industry names if they go over a certain character limit
  theme_minimal() +
  labs(
    title = "Top 10 Industries by Total Estimated Payroll",
    subtitle = "Clatsop County, Oregon (2025)",
    x = "Total Estimated Payroll",
    y = NULL,
    caption = "Source: BLS QCEW; Q1-Q3 2025"
  )

polished_plot

```

    
# 4. Brief Overview of Interactive Visualizations

You do not need to learn a completely new language to make interactive charts. The ***plotly*** package can automatically translate your ggplot into an interactive widget.

Notice that in the previous step, we saved our formatted chart to an object called polished_plot. Watch what happens when we feed it to ggplotly():

```{r}
# Install first if you haven't: install.packages("plotly")
library(plotly)

ggplotly(polished_plot)

```




# Day 1 Wrap-Up

You've made it! Today you learned how to:

    Navigate RStudio and manage your environment.

    Clean, filter, mutate, and reshape data using the tidyverse.

    Combine datasets using joins.

    Build, customize, and export professional visualizations using ggplot2.

Tomorrow: We move from local files to the cloud. We will connect directly to Census and BLS APIs, map spatial geographic data, and learn how to write dynamic, reproducible reports using Quarto.
