---
title: "Bonus: Cleaning Raw QCEW Data"
author:
  - "Tucker Plumlee and Paul Liu"
  - "R Training for LMI Shops: Beginning to Intermediate"
date: "June 2, 2026 — Bonus Material"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

The `oregon_counties_qcew.csv` file used throughout the Day 1 exercises did not come from BLS in that shape. It started as a folder of raw quarterly CSVs — one per geographic area — that required trimming, reshaping, and renaming before it became analysis-ready.

This bonus tutorial walks through how the file `oregon_counties_qcew.csv` was built. It is optional as nothing in Day 1 or Day 2 depends on it, but this tutorial can provide additional context on the data cleaning/wrangling process. The techniques here include reading many files at once (rather than one at a time), filtering to relevant rows, computing derived columns, and writing a clean output. The same workflow applies to most bulk public data — BLS LAUS, Census ACS Summary File tables, IPUMS state extracts, and similar sources that publish one file per geography per period.

The pipeline has five steps:

1. **Trim** the unzipped quarterly folder to the Oregon files only.
2. **Read** the Oregon files and stack them into one data frame.
3. **Filter** to the rows we care about — private sector, Q3, 6-digit NAICS.
4. **Compute** derived columns (average employment, annual wage) and rename for readability.
5. **Save** a clean, slim CSV.

---

# Getting the Raw Data

The data comes from the BLS Quarterly Census of Employment and Wages (QCEW): [https://www.bls.gov/cew/downloadable-data-files.htm](https://www.bls.gov/cew/downloadable-data-files.htm).

Navigate to **CSVs By Area** → **Quarterly** → **2025**. Download the ZIP for the relevant quarter and unzip it. The result is a folder of CSVs — one per geographic area (the U.S. as a whole, every state, every county, every MSA, plus other groupings). The Oregon county files have "Oregon" in their filenames; everything else covers other states or non-Oregon geographies.

Place the unzipped folder into your project — for this tutorial we will call it `Data/`.

**NOTE:** At the time this tutorial was created (May 2026), the most recent quarterly file is for the period Q1-Q3 2025.  

---

# Setup

Clear the workspace and load the tidyverse package:

```{r}
rm(list = ls())
library(tidyverse)
```

`rm(list = ls())` removes every object from the current session, which is useful at the top of a script that should run cleanly from scratch. Loading `tidyverse` brings in `dplyr`, `readr`, `stringr`, and `purrr` in one call.

Set the working directory and define a path to the data folder:

```{r}
setwd("C:/Users/Paul Liu/Desktop/R Course (June 2-3)")  # Use the file path to your own working directory!
data_path <- "Data"
```

Use forward slashes; Windows backslashes do not work in R paths.

---

# 1. Trim the Folder to Oregon Files Only

The unzipped folder contains hundreds of CSVs. We only need the Oregon county files. The Oregon files all contain `, Oregon` in their filenames, which makes filtering straightforward.

- `list.files()` returns a vector of file paths inside a folder.
- `grep()` matches a regular expression against a vector of strings. With `value = TRUE`, it returns the matching strings.

```{r}
all_files <- list.files(data_path, full.names = TRUE)  # Note that we defined the data_path object as simply "Data", the name of the subfolder where all the downloaded CSV files are stored.

oregon_files <- grep(", Oregon", all_files, value = TRUE) # Filters for all CSV files that contain the sub-string " ,Oregon"

length(all_files)
length(oregon_files)
```

`oregon_files` now holds the paths to the Oregon CSVs. Non-Oregon files remain on disk and will be ignored by the rest of this script.

### Exercises

1. Run `length(all_files)` and `length(oregon_files)`. What share of the original files did we keep?
2. Print the first few elements of `oregon_files`. Confirm each contains the `, Oregon` pattern.

---

# 2. Read and Combine the Oregon Files

Each Oregon county has its own CSV with identical column structure. To analyze them together, our approach is to **read each into a data frame** and **stack or append them into one combined data frame**.

The pattern:

1. Apply `read.csv()` to every file path, producing a list of data frames.
2. Use `bind_rows()` to stack the list into a single data frame.

```{r}
# Clean filenames (no path, no .csv extension) — optional
oregon_file_names <- tools::file_path_sans_ext(basename(oregon_files))

oregon_file_list <- lapply(oregon_files, read.csv)
names(oregon_file_list) <- oregon_file_names

oregon_counties <- bind_rows(oregon_file_list)
rm(oregon_file_list)  # free the memory
```

Two things to note:

- `lapply()` is base R's tool for applying a function to every element of a list or vector. The `map()` function the tidyverse equivalent of this.
- `bind_rows()` (from `dplyr`) requires compatible column structures across the data frames it stacks. Columns missing from any individual data frame are filled with `NA`, which accommodates files that are similar but not identical.

Passing `.id = "source_file"` to `bind_rows()` adds a column with the name of each source file, useful when you need to trace rows back to their origin. Not required here.

## Tidyverse Alternative

Steps 1 and 2 can be written as a single pipeline using `purrr::map()` and `bind_rows()`:

```{r}
oregon_counties <- list.files(data_path, pattern = ", Oregon", full.names = TRUE) %>%  # Filter for all Oregon files
  map(read.csv) %>%  # Read all selected CSV files into R
  bind_rows()   # Append all the files R just read in into one single data frame
```

Same result, but done in one short pipeline. The longer Base R version above is more transparent line-by-line; the short version is what you might see in production code.

### Exercises

1. Run `dim()` and `glimpse()` on `oregon_counties`. How many rows and columns?
2. Run `names(oregon_counties)`. The QCEW format has dozens of columns; which ones do you expect to keep?

---

# 3. Understanding the Raw QCEW Columns

Before filtering, it helps to understand what the rows mean. QCEW data describes employment and wages at every combination of:

- **Area** (`area_fips`, `area_title`) — a county, MSA, state, or other geography.
- **Industry** (`industry_code`, `industry_title`) — NAICS at varying levels of detail.
- **Ownership** (`own_code`) — private vs. various levels of government.
- **Aggregation level** (`agglvl_code`) — the level of summary.
- **Time** (`year`, `qtr`).

For the purpose of this course, we want to simplify the data by filtering just for a private-sector, county-level, and detailed-industry dataset for one quarter. We will therefore apply 3 filters:

- **`own_code == 5`** — private sector. Codes 1–4 are federal, state, local, and international government; codes 6–9 are various ownership combinations.
- **`qtr == 3`** — Q3 of 2025 (July–September). A full-year analysis would average or sum across quarters instead.
- **NAICS at the 6-digit level** — the most detailed industry codes. Excludes rolled-up sector and supersector codes, which would double-count if mixed with 6-digit codes.

For the NAICS filter we use a regular expression. The pattern `"\\d{6}$"` matches exactly six digits at the end of the string; `$` is the end-of-string anchor. `str_detect()` (from `stringr`) tests it.

See this cheat sheet on Regular Expressions if you want to learn more: https://rstudio.github.io/cheatsheets/regex.pdf

---

# 4. Filter, Compute, Rename, Select

A single pipeline handles the row filters, derived columns, renames, and final column selection.

```{r}
oregon_counties_clean <- oregon_counties %>%
  # Private sector, Q3, 6-digit NAICS
  filter(own_code == 5,
         qtr == 3,
         str_detect(as.character(industry_code), "\\d{6}$")) %>%

  # Derive/simplify the employment and wage columns
  mutate(
    avg_employment  = round((month1_emplvl + month2_emplvl + month3_emplvl) / 3),
    avg_annual_wage = avg_wkly_wage * 52,
    # Strip "NAICS 111110 " prefix (12 chars + space) from the industry title
    industry_title  = str_sub(industry_title, start = 13)
  ) %>%

  # Simplify column names
  rename(
    fips           = area_fips,
    naics          = industry_code,
    county         = area_title,
    industry_name  = industry_title,
    establishments = qtrly_estabs_count
  ) %>%

  # Keep only the columns we will use
  select(fips, county, naics, industry_name,
         establishments, avg_employment, avg_wkly_wage, avg_annual_wage)

glimpse(oregon_counties_clean)
length(unique(oregon_counties_clean$county))
```

Several choices in this pipeline are worth flagging:

- **`as.character()` inside `str_detect()`.** `industry_code` reads in as an integer by default; `str_detect()` requires a string. If `industry_code` were already character, the coercion would be a no-op.
- **`round()` on `avg_employment`.** Averaging three integer monthly headcounts produces decimals. Employment counts should not be fractional, so we round to the nearest whole.
- **`str_sub(industry_title, start = 13)`.** BLS industry titles arrive as `"NAICS 111110 Soybean farming"` — twelve characters of prefix, a space, then the title starting at position 13. The substring from position 13 onward removes the prefix. This assumes 6-digit NAICS throughout, which the filter guarantees; 4- or 5-digit NAICS would require a different start position.
- **`length(unique(...))`.** A sanity check on the number of distinct counties. Oregon has 36, plus the "Unknown Or Undefined" statewide rollup, for 37 total.

### Exercises

1. Swap `own_code == 5` for `own_code == 1` (federal government). How does the row count change? Which industries dominate?
2. Swap `qtr == 3` for `qtr == 1` and re-run. Does Oregon's industry mix differ meaningfully between Q1 and Q3?
3. Without rerunning the whole pipeline, write a `dplyr` chain that returns the top five industries statewide in `oregon_counties_clean` by total `avg_employment`.

---

# 5. Final Tidying Steps

Two final adjustments before saving:

```{r}
oregon_counties_clean <- oregon_counties_clean %>%
  mutate(
    county = str_remove(county, ", Oregon"),  # Can also be done with str_replace(county, ", Oregon", "")
    naics  = as.character(naics)
  )
```

- **Strip the state suffix.** `area_title` arrives as `"Baker County, Oregon"`. Once the data is filtered to Oregon, the suffix is redundant. The morning Wrangling tutorial assumes this cleanup has already happened.
- **Coerce NAICS to character.** NAICS codes are categorical rather than numeric — arithmetic on them would be meaningless — and codes with leading zeros (none in this data, but a habit worth keeping for other classification systems) would lose their leading digits if stored as integers.

---

# 6. Save the Cleaned CSV

```{r}
write_csv(
  oregon_counties_clean %>% select(-industry_name, -fips), # Get rid of industry_name and fips so we can do the join exercise later.
  "oregon_counties_qcew.csv"
  )
```

`industry_name` and `fips` are dropped on the way out. They get re-joined in the morning Wrangling tutorial — from `NAICS_crosswalk.csv` and `oregon_fips.csv` — to give students a motivating reason to learn `left_join()`. A fully self-contained file would keep both columns.

---

# Wrap-Up

The cleaning workflow covered:

- Bulk file management with `list.files()`, `grep()`, and (carefully) `file.remove()`.
- Reading many files with `lapply()` and stacking them with `bind_rows()`.
- Filtering rows on multiple conditions, including a regular expression.
- Computing derived columns and editing strings with `mutate()` and `stringr`.
- Renaming and selecting for a clean final shape.
- Writing the result to disk.

For deeper work in this area:

- **`purrr`** provides tidyverse equivalents to `lapply()` and friends.
- **`fs`** is a tidyverse alternative to base R file operations (`fs::dir_ls()`, `fs::file_delete()`) with safer defaults.
- **`vroom`** reads CSVs faster than `read.csv()` or `read_csv()` on large files.

The patterns transfer directly to other bulk public data: BLS LAUS county files, Census ACS Summary File tables, IPUMS state-by-state extracts, and similar sources.

For continued practice: [R for Data Science](https://r4ds.hadley.nz/), particularly the iteration chapter on `purrr::map()`, and the [purrr cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf).
