---
title: "Digging Deeper: More R Features and Workflows"
date: 2026-06-03T09:30
date-format: "dddd, MMMM D, YYYY; h:mmA"
---
# Setup and Introduction {#sec-introduction}

As usual, let's start by loading the various `tidyverse` libraries.

```{r}
#| label: setup
#| message: false

library(tidyverse)

# Remember to check your working directory using getwd() and set it to your training materials folder (the one containing the Data and Tutorials sub-folders) if it isn't already using setwd()
```

In the R Fundamentals and Data Wrangling sessions, we learned enough to load data, reshape it, and produce summaries. This session reviews and fills in the some of the conceptual foundation underneath that work as well as introduces tools that will become essential as your scripts grow more complex.

We will use the same Oregon QCEW data from earlier sessions:

```{r}
#| label: load-data
#| message: false

oregon_qcew <- read_csv("Data/oregon_counties_qcew.csv")
```

# Data Types {#sec-data-types}

Every value in R has a **type** that determines what you can do with it. This is similar to data types in Excel--if your data is the wrong type for the function you ar trying to use, it won't work (or will produce unexpected results). Unlike Excel, data types in R are implied by specific syntax. Knowing the type of your data helps you avoid errors and understand why certain operations work or fail.

## The Core Types

R has four types you will encounter constantly:

```{r}
#| label: data-types

# Character (text/strings) — always surrounded by quotes
county_name <- "Multnomah County"
class(county_name)

# Integer — whole numbers, created with an L suffix
employee_count <- 250L
class(employee_count)

# Numeric (double/float) — numbers with or without decimals
wage <- 52000.50
class(wage)

# Logical — TRUE or FALSE (the result of any comparison)
is_large <- wage > 50000
is_large
class(is_large)
```

You can check a value's type with `class()`, or test for a specific type with the `is.` family of functions:

```{r}
#| label: type-checking

is.character(county_name)
is.numeric(wage)
is.logical(is_large)
```

## Why Types Matter

Types determine what operations make sense. You can add two numbers but not two character strings:

```{r}
#| label: type-errors
#| error: true

# This works
10 + 20

# This does not
"10" + "20"
```

The value `"10"` looks like a number but it is a character string. R treats it as text and has no idea how to add text together with `+`. This kind of type mismatch is one of the most common sources of errors when reading in data --- a column you expect to be numeric might have been read as character because of a stray comma or footnote symbol somewhere in the file.

## Converting Between Types

When types don't match, you can convert with the `as.` functions:

```{r}
#| label: type-conversion

# Character to numeric
as.numeric("42")

# Numeric to character
as.character(42)

# Logical to numeric (TRUE becomes 1, FALSE becomes 0)
as.numeric(TRUE)
sum(c(TRUE, FALSE, TRUE, TRUE))   # Counting TRUEs
```

A particularly useful function from the readr package (loaded with tidyverse) is `parse_number()`, which pulls a number out of messy text:

```{r}
#| label: parse-number

parse_number("$52,000.50")
parse_number("approximately 1,200 employees")
```

This is handy when importing data where numeric columns contain dollar signs, commas, or other formatting characters. Keep in mind that that there needs to be numeric values in the text---the function cannot parse number names (i.e., numbers in word form).

```{r}
#| label: parse-number-error
#| error: true

parse_number("fifty")
```

::: {.callout-tip collapse="true"}
## Exercise 1: Types and Conversion

**Part A:** What type is each of these values? Predict first, then check with `class()`:

```{r}
#| label: exercise-1a
#| eval: false

class(3.14)
class(100L)
class("TRUE")
class(TRUE)
class(NA)
```

**Part B:** What happens when you run `as.numeric("hello")`? Why?

```{r}
#| label: exercise-1b
#| eval: false

as.numeric("hello")
```
:::

# Data Structures {#sec-data-structs}

Data types describe individual values. **Data structures** describe how values are organized together. You have already worked with vectors and data frames; let's formalize what they are and add lists to the picture.

## Vectors

A **vector** is a sequence of values that are all the **same type**. You create one with `c()`:

```{r}
#| label: vectors

wages <- c(45000, 52000, 38000, 61000)
counties <- c("Baker", "Benton", "Clackamas", "Clatsop")
above_50k <- c(FALSE, TRUE, FALSE, TRUE)
```

Vectors are the most fundamental structure in R. Even a single value like `wage <- 52000` is actually a vector of length 1. Every column in a data frame is a vector.

If you mix types in a vector, R silently **coerces** everything to the most flexible type. This is like an order of operations where logical → integer → numeric → character:

```{r}
#| label: coercion

mixed <- c(1, 2, "three")
mixed
class(mixed)
```

Notice that the numbers `1` and `2` were quietly converted to character strings. This is a common source of subtle bugs.

## Lists

A **list** is like a vector that *can* hold different types, including other lists, data frames, or even model outputs (e.g., coefficients, residuals, test statistics, etc.):

```{r}
#| label: lists

my_list <- list(
  name   = "Oregon",
  pop    = 4240137,
  counties = c("Baker", "Benton", "Clackamas"),
  is_state = TRUE
)

my_list
```

Notice that each element in the list above is named (i.e., `name = value`), much like you would assign a vector a name in order to store it as an object for future use. Similarly, names can help with retrieving a list's elements for later use (we'll see how in the next section on indexing).

Lists are the Swiss Army knife of R data structures. Many R functions return lists, and you will often encounter them when working with model outputs, API responses, or nested data.

## Data Frames (and Tibbles)

A **data frame** is just a *list* of named vectors that all have the same length (i.e., the same number of elements). This is the table structure you are already familiar with, where each named vector is a column/field, and each row (or the same position within each vector) is an observation. You can create your own data frame from scratch using the `data.frame()` function (much like you would create a new list with `list()`):

```{r}
#| label: dataframe

my_df <- data.frame(
  county = c("Baker", "Benton", "Clackamas"),
  employment = c(5000, 45000, 180000),
  wage = c(42000, 55000, 62000)
)

my_df
```

Notice that the result of printing/displaying this data frame object is a bit different than what we've seen throughout the course. This is because `tidyverse` packages like `readr`, `tidyr`, or `dplyr` create and work with a special kind data frame structured called a **`tibble`**. The differences between the two are described below (@sec-tidyverse-base).

::: {.callout-tip collapse="true"}
## Exercise 2: Structures

**Part A:** Create a list that contains your name (character), your age (numeric), and a vector of three hobbies (character vector). Print it.

**Part B:** Create a vector of five city names. Then create a vector of five population numbers. Combine them into a data frame named `city_data` using `data.frame()`. Your results should look similar to the output below.

```{r}
#| label: exercise-2
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

cities <- c("Portland", "Salem", "Eugene", "Gresham", "Hillsboro")
populations <- c(640623, 179977, 179352, 115739, 112035)
city_data <- data.frame(city = cities, pop = populations)
city_data
```
:::

# Indexing and Subsetting {#sec-indexing}

**Indexing** is how you pull out specific pieces of a data structure. Even though tidyverse functions like `filter()` and `select()` handle most of this for data frames, this is a fundamental operation you will need when working with vectors, lists, and in programming contexts like loops and functions (which we'll cover later today, time allowing).

## Indexing Vectors with `[]`

Use square brackets with a position number to retrieve a value at that position number:

```{r}
#| label: vector-indexing

wages <- c(45000, 52000, 38000, 61000)

# Single element
wages[1]
wages[3]

# Multiple elements
wages[c(1, 3)]

# All except the second element
wages[-2]

# Logical indexing: elements where a condition is TRUE
wages[wages > 40000]
```

## Indexing Lists with `[]` and `[[]]`

Lists use two kinds of brackets for indexing:

```{r}
#| label: list-indexing

my_list <- list(name = "Oregon", pop = 4240137, counties = c("Baker", "Benton"))

# Single brackets return a *sub-list* of values at the specified position
my_list[1]
class(my_list[1])

# Double brackets extract the *element itself*
my_list[[1]]
class(my_list[[1]])

# You can also access named elements with double brackets
my_list[["pop"]]

# Use a combination of the two to access sub-elements of list elements
my_list[[3]][2]
```

Think of a list like a train: `[1]` gives you the first train car (still a train car), while `[[1]]` gives you the first row *inside* the first car.

## Indexing Data Frames with `$` and `[]`

Data frames support several indexing styles:

```{r}
#| label: df-indexing

# The $ operator extracts a single column as a vector
my_df$county

# Since data frames are lists, you can use $ to similarly access named list elements
my_list$pop

# This is the most common way to grab a column outside of a tidyverse pipeline
mean(oregon_qcew$avg_annual_wage)
```

You can also use bracket notation with `[row, column]`:

```{r}
#| label: df-bracket-indexing

# Single cell: row 1, column 3
oregon_qcew[1, 3]

# Entire row 1
oregon_qcew[1, ]

# Entire column by name
oregon_qcew[, "avg_employment"] |> head()

# First 5 rows of two columns
oregon_qcew[1:5, c("county", "avg_employment")]
```

The `$` syntax is especially useful for quick calculations and when passing a column to a function that expects a plain vector rather than a data frame:

```{r}
#| label: dollar-uses

# Calculate the range of annual wages
max(oregon_qcew$avg_annual_wage) - min(oregon_qcew$avg_annual_wage)

# Use in a base R function
table(oregon_qcew$county)[1:5]
```

::: {.callout-tip collapse="true"}
## Exercise 3: Indexing Practice

```{r}
#| label: exercise-3-setup

fruits <- c("apple", "banana", "cherry", "date", "elderberry")
info <- list(fruit = "apple", color = "red", seeds = 5)
```

**Part A:** Extract the third element of `fruits`. Then extract the first and fifth elements together.
```{r}
#| label: exercise-3-a
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

# Part A
fruits[3]
fruits[c(1,5)]
```

**Part B:** Extract the `color` element from `info` using `$` notation. Then extract it using `[[]]` notation.
```{r}
#| label: exercise-3-b
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

# Part B
info$color
info[["color"]]
```

**Part C:** Use `$` to find the `mean()` of the `establishments` column in `oregon_qcew`.
```{r}
#| label: exercise-3-c
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

mean(oregon_qcew$establishments)
```
:::

# Handling Missing Values (NA) {#sec-na}

Real data almost always has gaps (i.e., empty cells in a spreadsheet). In R, missing values are represented by `NA` (Not Available). Understanding how `NA` behaves is critical because a single missing value can silently derail your calculations.

## How NA Works

`NA` is contagious --- any operation involving an `NA` returns `NA`:

```{r}
#| label: na-contagious

x <- c(10, 20, NA, 40)

sum(x)
mean(x)
```

R is being cautious: it does not know what the missing value represents, so it cannot give you a definitive sum or mean. You saw `na.rm = TRUE` in passing during earlier sessions --- this tells R that it should "remove" or ignore `NA` values when executing the function over the specified values.

```{r}
#| label: na-rm

sum(x, na.rm = TRUE)
mean(x, na.rm = TRUE)
```

## Detecting Missing Values

You cannot test for `NA` with `==` because `NA == NA` returns `NA` (the logic being "I don't know if this unknown thing equals that unknown thing"). Use `is.na()` instead:

```{r}
#| label: detect-na

# Returns a logical vector IDing if each value is NA
is.na(x)

# Since TRUE==1, you can use this to count missing values in a column
sum(is.na(oregon_qcew$avg_annual_wage))
```

## Removing and Replacing NAs in Pipelines

The tidyverse provides some useful tools for working with NAs inside data pipelines:

```{r}
#| label: na-tidyverse

# replace_na() substitutes a value for NA
x_filled <- replace_na(x, 0)
x_filled

# fill() replaces NA values with the previous/next non-NA value in that column
city_pops <- tibble(
  city = c("Portland", "Eugene", "Portland", "Salem", "Eugene"),
  year = c(2020, NA, 2025, NA, NA),
  pop = c(652498, 176759, 640623, 179977, 179352)
  )

city_pops

city_pops |> 
  fill(year, .direction = "down")

# drop_na() removes rows with NA in specified columns
city_pops |>
  drop_na(year)

# We don't have to worry about any NA values in our QCEW data because suppressions were already replaced with 0's)

nrow(oregon_qcew)

oregon_qcew |>
  drop_na(avg_annual_wage) |>
  nrow()
```

::: {.callout-tip collapse="true"}
## Exercise 4: Working with NAs

Create a vector `temps <- c(72, 68, NA, 75, NA, 80)`. Calculate the mean, handling the NAs properly. Then count how many NAs are in the vector.

```{r}
#| label: exercise-4
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

temps <- c(72, 68, NA, 75, NA, 80)
mean(temps, na.rm = TRUE)
sum(is.na(temps))
```
:::

# Tidyverse vs. Base R {#sec-tidyverse-base}

So far we have been learning the **tidyverse** way of doing things, but it is worth understanding the older, built-in way of doing many of the same tasks in what is called **base R**. This includes all of the "built-in" functions that are provided when you install R from [The R Project website](https://www.r-project.org/) (e.g., `data.frame()`, `sum()`, `mean()`, etc.).

Despite arguments on the internet, neither is "better" --- they are two dialects of the same language, and you will encounter both in the wild.

## Philosophy

As the foundations of the language, base R came first and uses a more general-purpose programming style. The tidyverse---initially developed by Hadley Wickham (and others) starting around 2014--- was designed specifically for data analysis and emphasizes readable, composable pipelines.

| Concept | Base R | Tidyverse |
|---------|--------|-----------|
| Subset rows | `df[df$col > 5, ]` | `df |> filter(col > 5)` |
| Select columns | `df[, c("a", "b")]` | `df |> select(a, b)` |
| Create columns | `df$new <- df$x + 1` | `df |> mutate(new = x + 1)` |
| Group summaries | `aggregate(x ~ group, df, mean)` | `df |> group_by(group) |> summarize(mean(x))` |
| Read CSV | `read.csv("file.csv")` | `read_csv("file.csv")` |

The tidyverse version tends to be more readable for data tasks because the pipe lets you read the steps top to bottom, like a recipe. Base R indexing is more compact and does not require loading packages.

## Tibbles vs. Data Frames

A **tibble** is the tidyverse's version of a data frame. When you use `read_csv()`, you get a tibble. It behaves almost identically to a base R data frame, with a few differences:

```{r}
#| label: tibble-vs-df

# Base R data frame
base_df <- data.frame(
  name = c("Alice", "Bob", "Carol"),
  score = c(90, 85, 92)
)

# Tibble
tidy_tb <- tibble(
  name = c("Alice", "Bob", "Carol"),
  score = c(90, 85, 92)
)

# Printing: tibbles show type information and truncate neatly
tidy_tb
base_df
```

Key differences: tibbles print more informatively (showing types and truncating large datasets instead of flooding your console), they never silently convert strings to factors (another data type we haven't reviewed in this training, but is worth understanding), and they give stricter error messages when you try to access a column that does not exist. In practice, tibbles and data frames are interchangeable for most tasks---any tidyverse functions will work on base R data frames as well as tibbles, and vice-versa.

## The Two Pipes: `%>%` and `|>`

The Data Wrangling session used `%>%`, the "pipe" operator from the `magrittr` package, which is loaded as part of the tidyverse. For some time, this was the only pipe operator available in R. However, base R version 4.1+ added a **native pipe** `|>` that is built into R and does not require any package:

```{r}
#| label: pipes

# magrittr pipe (requires tidyverse or magrittr)
oregon_qcew %>% nrow()

# Native pipe (built into R, no packages needed)
oregon_qcew |> nrow()
```

They work the same way for most tasks. The Day 2 walkthroughs for this training use `|>` because it is the direction R is heading and does not depend on any package. You will see `%>%` in older code and tutorials --- both are fine.

# More mutate() Tricks {#sec-mutate-tricks}

The Data Wrangling session introduced the basics of using the `mutate` function (from the `dplyr` tidyverse package) to create new columns. It is worth highlighting some other useful operations that can be performed with `mutate`. As with most things in R (and programming in general) there can be multiple ways to accomplish the same task.

## Removing Variables Inside mutate()

In the Data Wrangling session you learned to drop columns with `select(-col)`. You can also remove a column inside `mutate()` by setting it to `NULL`. This is useful when you want to create new columns and clean up old ones in the same step:

```{r}
#| label: mutate-remove

oregon_qcew |>
  mutate(
    wage_thousands = avg_annual_wage / 1000,
    avg_annual_wage = NULL
  ) |>
  select(county, naics, wage_thousands) |>
  head()
```

The original `avg_annual_wage` column is gone, and `wage_thousands` takes its place. You can also use this to remove helper columns you created mid-pipeline.

## Applying Functions Across Multiple Columns with `across()`

Often you need to do the same operation on several columns at once. Instead of writing out each one, use `across()`:

```{r}
#| label: across

# Round multiple numeric columns at once
oregon_qcew |>
  mutate(across(c(avg_wkly_wage, avg_annual_wage), round)) |>
  select(county, avg_wkly_wage, avg_annual_wage) |>
  tail()

# Round multiple numeric columns to the hundredths place
oregon_qcew |>
  mutate(across(c(avg_wkly_wage, avg_annual_wage), \(x) round(x, -2))) |>
  select(county, avg_wkly_wage, avg_annual_wage) |>
  tail()
```

The first argument to `across()` is a selection of columns (you can use the same helpers as in `select()`, such as `starts_with()` or `where(is.numeric)`). The second argument is the function (or list of functions) to apply. You can specify a function in a few different ways:

- Simply providing its name (if no other arguments are required aside from the column/vector being transformed)
- Using an *anonymous function* where the column/vector being transformed is specified as a variable using `\(x)` followed by the function with its necessary arguments: `round(x, -2)`.
- Using a *lambda function* where the function comes after a `~` and the column/vector being transformed is represented as `.x`.

The `across()` function can also be used in `summarize()` to apply a single aggregation function to multiple columns at once:

```{r}
# Summarize multiple columns in one step
oregon_qcew |>
  summarize(
    across(
      c(establishments, avg_employment),
      mean
      )
  )
  
# Summarize multiple columns using multiple functions in a list
oregon_qcew |>
  summarize(
    across(
      c(establishments, avg_employment),
      list(total = sum, avg = mean)
      )
  )
```

If a named list of functions is provided to across, the resulting columns will be named by combining the column name and function name with an underscore in between (`{.col}_{.fun}`).

::: {.callout-tip collapse="true"}
## Exercise 5: mutate() and across()

**Part A:** Using the `oregon_qcew` data, create a new column `emp_per_estab` (employment per establishment), then remove the `establishments` column, all inside one `mutate()` call.

```{r}
#| label: exercise-5a
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

oregon_qcew |>
  mutate(
    emp_per_estab = avg_employment/establishments,
    establishments = NULL
  ) |>
  tail()
```

**Part B:** Use `summarize()` with `across()` and `where(is.numeric)` to calculate the `mean` of every numeric column in `oregon_qcew`.

```{r}
#| label: exercise-5b
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

oregon_qcew |>
  summarize(across(where(is.numeric), mean))
```
:::

## Working with Strings Using stringr

Text data is messy: extra spaces, inconsistent capitalization, codes embedded in names. The `stringr` package (loaded with tidyverse) provides a consistent set of functions for working with strings. They all start with `str_`, making them easy to find and remember.

### Detecting Patterns

`str_detect()` returns `TRUE` or `FALSE` for whether a pattern appears in a string. This is extremely useful inside `filter()`:

```{r}
#| label: str-detect

# Find all rows for Lane County
oregon_qcew |>
  filter(str_detect(county, "Lane")) |>
  head()
```

### Replacing and Removing Text

We can easily remove the repetivite "County" suffix from all of the county names using `str_remove`, or replace it with "Co." using `str_replace`. 

```{r}
#| label: str-replace

# Remove a consistent suffix from county names
oregon_qcew |>
  mutate(county_short = str_remove(county, " County")) |>
  select(county, county_short) |>
  distinct() |>
  head()

# Replace text
oregon_qcew |>
  mutate(county_short = str_replace(county, " County", " Co.")) |>
  select(county, county_short) |>
  distinct() |>
  head()

# We could also use str_replace to remove "County" by replacing it with an empty string ""
oregon_qcew |>
  mutate(county_short = str_replace(county, " County", "")) |>
  select(county, county_short) |>
  distinct() |>
  head()
```

### Changing Case

```{r}
#| label: str-case

str_to_upper("multnomah county")
str_to_lower("MULTNOMAH COUNTY")
str_to_title("multnomah county, oregon")
```

### Subsetting and Splitting

Like with LEFT(), RIGHT() and MID() in Excel, you can extract a sub-string using certain indices/positions in the string using `str_sub()`. You can also split a string using a specific character (note that the output will be a list with multiple elements).

```{r}
#| label: str-extract

# Extract a number from text
str_sub("NAICS 3254: Pharmaceutical", start = 7, end = 10)

# Split a string
str_split("Baker County, Oregon", ", ")
```

### Padding and Trimming

```{r}
#| label: str-pad-trim

# Pad FIPS codes to consistent width (common with county codes)
str_pad(c("1", "23", "456"), width = 3, pad = "0")

# Trim whitespace from both ends
str_trim("  extra spaces  ")
str_squish("  too   many    spaces  ")
```

::: {.callout-tip collapse="true"}
## Exercise 6: stringr Practice

**Part A:** Use `mutate()` and `str_remove()` to create a column called `county_clean` that removes `", County"` from the `county` column. Then use `str_to_upper()` to create an all-caps version of `county_clean`.

```{r}
#| label: exercise-6a
#| code-fold: true
#| code-summary: "Click here to reveal the solution"
#| 
oregon_qcew |>
  mutate(
    county_clean = str_remove(county, " County"),
    county_upper = str_to_upper(county_clean)
  ) |>
  select(county, county_clean, county_upper) |>
  distinct()
```

**Part B:** Filter the data to only rows where the county name contains the word "Washington". (Careful --- does any county name contain "Washington" as part of a longer word?)

```{r}
#| label: exercise-6b
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

oregon_qcew |>
  filter(str_detect(county, "Washington"))
```
:::

# Additional Essential Tidyverse Functions {sec-tidyverse-essentials}

While the Data Wrangling session provided the absolute essential functions you'll need when using tidyverse to clean data, there are a few others that fill common gaps between these core verbs.

## `count()`: Quick Frequency Tables

`count()` is a shortcut for `group_by() |> summarize(n = n())`. It counts how many rows exist for each value of a variable:

```{r}
#| label: count

# How many industries per county?
oregon_qcew |>
  count(county, sort = TRUE) |>
  head()
```

Notice that the output automatically adds a column named `n` with the frequencies. Add `sort = TRUE` to automatically arrange the results from most to fewest. You can also count combinations of multiple variables

```{r}
#| label: count-multi
#| eval: false

# If the data had a year column, you could count by county and year:
oregon_qcew |> count(county, year)
```

## `distinct()`: Unique Rows

`distinct()` returns only the unique values of specified columns. It is the tidyverse equivalent of `unique()`:

```{r}
#| label: distinct

# List of unique counties
oregon_qcew |>
  distinct(county) |>
  head()

# Count unique counties
oregon_qcew |>
  distinct(county) |>
  nrow()
```

By default, `distinct()` drops all columns not named as arguments. Add `.keep_all = TRUE` to keep the full row (it keeps the first occurrence of each distinct value):

```{r}
#| label: distinct-keep-all

oregon_qcew |>
  distinct(county, .keep_all = TRUE) |>
  head()
```

## `slice()` and Friends: Rows by Position

While `filter()` selects rows by condition, the `slice()` family selects rows by position or rank:

```{r}
#| label: slice

# First 3 rows
oregon_qcew |> slice(1:3)

# Last 3 rows
oregon_qcew |> slice_tail(n = 3)

# Row with the highest avg_annual_wage
oregon_qcew |> slice_max(avg_annual_wage, n = 1)

# Row with the lowest avg_annual_wage
oregon_qcew |> slice_min(avg_annual_wage, n = 1)

# Random sample of 5 rows
oregon_qcew |> slice_sample(n = 5)
```

`slice_max()` and `slice_min()` are especially useful combined with `group_by()` --- for example, finding the highest-paying industry in each county:

```{r}
#| label: slice-grouped

oregon_qcew |>
  group_by(county) |>
  slice_max(avg_annual_wage, n = 1) |>
  ungroup() |>
  select(county, naics, avg_annual_wage) |>
  head(10)
```

## `pull()`: Extract a Column as a Vector

`pull()` extracts a single column from a data frame as a plain vector. This is the pipe-friendly equivalent of `$`:

```{r}
#| label: pull

# These two are equivalent:
oregon_qcew |> pull(avg_annual_wage) |> mean()
mean(oregon_qcew$avg_annual_wage)
```

`pull()` is useful at the end of a pipeline when you need a vector for a function that does not accept a data frame (e.g., `mean`).


::: {.callout-tip collapse="true"}
## Exercise 7: Putting It Together

Using the `oregon_qcew` data:

**Part A:** Use `count()` and `slice_max()` to find the 5 counties with the most industry records.
```{r}
#| label: exercise-7a
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

oregon_qcew |>
  count(county) |>
  slice_max(n, n = 5)
```

**Part B:** Use `slice_max()` after a `group_by()` to find the highest-employment industry in each of the first 5 counties (alphabetically). Hint: you'll want to filter to the first 5 counties first, either by explicitly filtering based on their names or by programmatically getting the list of names using `distinct()`, `arrange()`, and `slice()` (or some similar approach).

```{r}
#| label: exercise-7b
#| code-fold: true
#| code-summary: "Click here to reveal the solution"

# Create a new data frame with just the top 5 distinct county names after sorting alphabetically
first_alpha_counties <- oregon_qcew |>
  distinct(county) |> 
  arrange(county) |> 
  slice(1:5)

oregon_qcew |> 
  filter(county %in% first_alpha_counties$county) |> 
  group_by(county) |> 
  slice_max(order_by = avg_employment, n = 1) |> 
  select(county, naics, avg_employment)
```
:::

## Wrapping Up

This session covered the conceptual building blocks that support everything else you do in R:

-   **Data types** (character, numeric, integer, logical) determine what operations make sense. Use `class()` to check and `as.` functions to convert.
-   **Data structures** (vectors, lists, data frames) organize values. Vectors hold same-type elements, lists hold anything, and data frames are the table structure for analysis.
-   **Indexing** with `[]`, `[[]]`, and `$` lets you extract specific pieces from any structure.
-   **NA values** require deliberate handling with `na.rm`, `is.na()`, `drop_na()`, and `replace_na()`.
-   **Factors** give you control over categorical data ordering, especially important for charts.
-   **Tibbles** are the tidyverse's version of data frames; `|>` is R's native pipe alongside tidyverse's `%>%`.
-   **stringr** functions (`str_detect`, `str_remove`, `str_replace`, `str_to_upper`, `str_pad`) make text manipulation clean and consistent.
-   **count()**, **distinct()**, **slice()**, **pull()**, **across()**, and removing columns with `mutate(col = NULL)` round out your tidyverse toolkit.

These concepts become more important as your scripts grow from simple summaries to full analytical pipelines. The more comfortable you are with types, structures, and indexing, the faster you will be able to diagnose problems when something goes wrong.