---
title: Accessing Public Data via APIs
date: 2026-06-03T13:00
date-format: "dddd, MMMM D, YYYY; h:mmA"
execute: 
  message: false
---

# API Basics

## What is an API? Why use one?
In general, an **API** (Application Programming Interface) is a way for two software systems to talk to each other. In the context of data, an API provides a structured way to request specific data from an organization's servers and get it back in a predictable format. 

Without an API, retrieving data from the Census Bureau or the Bureau of Labor Statistics typically means navigating a website, clicking through a data tool, selecting your options, and downloading a CSV or Excel file. This process is manual, time-consuming, and hard to reproduce. If the data updates next quarter, you have to do it all over again.

With an API, you write code that sends a request and receives data back directly into your R environment. This means your entire data retrieval process is scripted, reproducible, and can be re-run whenever the data is updated.

## What does an API look like?

API providers publish **documentation** that describes what data is available, how to request it, and what format the response takes.

As an example, let's take a look at the documentation for the Bureau of Labor Statistics' API: <https://www.bls.gov/developers/api_signature_v2.htm#multiple>

There's a lot of jargon here that, for most of you, is probably pretty foreign: HTTP types, header, payload, JSON... 

## API "Wrapper" Packages
Obviously, if you aren't already fluent in the language of internet protocols, this can present a pretty high barrier for using an API to access the data you want. This is where **wrapper packages** come in.

Thankfully, there is a kind and helpful community of developers out there who have written packages for R (and Python, etc.) that "wrap" all the complexity of constructing API requests, sending them, and parsing the responses in functions that look more familiar and are easier to use. You get tidy data frames back, ready for analysis with dplyr, ggplot2, and the rest of the tidyverse.

Today we will work with two wrapper packages:

-   **tidycensus** --- for accessing U.S. Census Bureau data
-   **blsAPI** --- for accessing Bureau of Labor Statistics data 

If we have time, we'll also look at one provided by IPUMS, which integrates microdata from surveys such as the Current Population Survey (CPS) and American Community Survey (ACS).

# `tidycensus`

## Overview

The [tidycensus](https://walker-data.com/tidycensus/index.html) package is a "wrapper" around the Census API for a variety of datasets including the decennial census, ACS, and population estimates. -Kyle Walker, one of the developers of the package, has also written a book, [*Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r/) that is available for free online: <https://walker-data.com/census-r/>. 

## Getting Started

We'll begin by ensuring that `tidycensus` is installed and loaded into our environment.

```{r}
#| label: setup

# install.packages('tidycensus')
library(tidycensus)

# Also make sure we have tidyverse if it isn't already loaded
library(tidyverse)
```

Now, we need to save our API key somewhere that the package can easily access it without us needing to copy and paste it every time we make a request. Luckily, `tidycensus` makes this easy for us by providing a function that will automatically save it to our `.Renviron` file (this file stores many values that R or certain packages may need to regularly access behind the scenes).

```{r}
#| label: census-key
#| eval: false

# Replace "YOUR_KEY" with the key you received from Census
census_api_key("YOUR_KEY", install = TRUE)

# Reload R environment 
readRenviron("~/.Renviron")

# You can check it with:
Sys.getenv("CENSUS_API_KEY")
```

::: {.callout-important}
## Security Best Practice
Never hardcode your API key directly in a script that you share or commit to version control. Using `census_api_key(install = TRUE)` stores the key in your personal `.Renviron` file, keeping it out of your code.
:::

## Retrieving ACS Data for Oregon Counties

Let's walk through a practical example: comparing population change across Oregon counties using the American Community Survey, specifically the 5-year estimates.

### Step 1: Get Total Population by County

The `get_acs()` function retrieves data from the American Community Survey. We need to specify a few things: the geography level, the state, the survey year, and the variable(s) we want. ACS provides data for thousands of variables with unique codes. Thankfully, `tidycensus` provides a helpful function for browsing variables (for ACS and many other Census datasets): `load_variables()`

```{r}
#| label: search-variables
#| eval: false

# Browse the available ACS 5-year variables for 2024
# cache stores the results so we don't have to wait for them to load every time
vars_2024 <- load_variables(2024, "acs5", cache = TRUE)
View(vars_2024)
```

With some searching we can see that the variable code for total population is **B01003_001**. 

With variable code in hand, let's pull total population for Oregon counties from two time periods. We'll start with the 5-year estimates from 2019.

```{r}
#| label: acs-2019

# 2015-2019 5-year ACS estimates
or_pop_2019 <- get_acs(
  geography = "county",
  variables = "B01003_001",
  state = "OR",
  year = 2019,
  geometry = FALSE
)

head(or_pop_2019)
```

Notice the structure: each row represents one county, with the `estimate` column containing the population value and `moe` containing the margin of error.

When pulling data for 2024 we will also set `geometry = TRUE` to download the county shapefiles provided through Census' TIGER/Line program. 

```{r}
#| label: acs-2024

# 2020-2024 5-year ACS estimates
or_pop_2024 <- get_acs(
  geography = "county",
  variables = "B01003_001",
  state = "OR",
  year = 2024,
  geometry = TRUE
)

head(or_pop_2024)
```

Notice that our view of the data frame includes a bunch of extra information related to the geometric/shapefile data. We will not use the shapefiles now, but they will be useful in a later mapping session. We're going to do a bit of additional work before saving the dataset to use later. 

### Step 2: Calculate Population Growth

Let's combine the two datasets and calculate the percentage change in population:

```{r}
#| label: pop-growth

# Prepare each dataset: keep only GEOID, NAME, and estimate; rename estimate
or_pop_2019 <- or_pop_2019 |>
  select(GEOID, NAME, estimate) |>
  rename(pop_2019 = estimate)

or_pop_2024 <- or_pop_2024 |>
  select(GEOID, NAME, estimate) |>
  rename(pop_2024 = estimate)

# Join and calculate percent change
or_pop_change <- or_pop_2024 |>
  left_join(or_pop_2019, by = c("GEOID", "NAME")) |>
  mutate(
    pop_change = pop_2024 - pop_2019,
    pct_change = (pop_change / pop_2019) * 100
  ) |>
  # Clean up county names (remove ", Oregon" suffix)
  mutate(county = str_remove(NAME, " County, Oregon"))

head(or_pop_change)
```

Now that we have a combined dataset, let's save it to our Data folder using the `.RDS` format so we can use the shapefile data for mapping later. After saving the data, we'll remove the geometry data from the object we are going to work with in this session.

```{r}
#| label: geo-data-save

saveRDS(or_pop_change, "Data/or_geo_data.RDS")

# Convert the data to a normal tibble/data frame and remove the geometry field
or_pop_change <- or_pop_change |>
  tibble() |>
  select(-geometry)
```

### Step 3: Visualize Population Growth

Let's create a bar chart showing population growth by county, sorted from highest to lowest growth:

```{r}
#| label: fig-pop-growth
#| fig-cap: "Percentage Population Change in Oregon Counties (2019-2024)"
#| fig-alt: "Horizontal bar chart showing population percent change for Oregon counties, sorted in descending order."
#| fig-width: 8
#| fig-height: 10
#| fig-align: center

or_pop_change |>
  ggplot(aes(x = pct_change, y = reorder(county, pct_change))) +
  geom_col(aes(fill = pct_change > 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "red3")) +
  labs(
    x = "Population Change (%)",
    y = NULL,
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
## Exercise 1: Explore a Different Variable

Using `load_variables()`, find the variable code for **median household income** (hint: look in table B19013). Then use `get_acs()` to retrieve median household income for Oregon counties from the 2024 5-year ACS. Create a horizontal bar chart showing the top 20 counties with the highest median income.

```{r}
#| label: exercise-1
#| eval: false

# Step 1: Search for the variable
vars <- load_variables(2024, "acs5", cache = TRUE)

# Filter for table B19013 to find median household income
# vars |> filter(str_detect(name, "B19013"))

# Step 2: Retrieve the data
or_income <- get_acs(
  geography = "county",
  variables = "B19013_001",
  state = "OR",
  year = 2024
) |>
  mutate(county = str_remove(NAME, " County, Oregon"))

# Step 3: Create a bar chart
or_income |>
  slice(n = 20) |>
  ggplot(aes(x = estimate, y = reorder(county, estimate))) + 
  geom_col(fill = "steelblue") +
  labs(
    title = "Median Household Income in Oregon Counties, 2024",
        x = "Median Income ($)",
    y = NULL,
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates"
  )
```
:::

## Other Available Datasets

While we focused on the ACS today, tidycensus provides access to several other Census Bureau datasets:

-   **Decennial Census** via `get_decennial()` --- full-count data from the 2000, 2010, and 2020 Census
-   **Population Estimates** via `get_estimates()` --- annual population estimates, including components of change (births, deaths, migration)
-   **Migration Flows** via `get_flows()` --- county-to-county and metro-to-metro migration data from the ACS

::: {.callout-tip collapse="true"}
## Exercise 2: Try Another Dataset

Use `get_decennial()` to retrieve total population by county in Oregon from the 2020 Decennial Census. The variable for total population in the DHC summary file is **P1_001N**. How do the Decennial counts compare to the ACS estimates you retrieved earlier?

```{r}
#| label: exercise-2
#| eval: false

or_decennial_2020 <- get_decennial(
  geography = "county",
  variables = "______",
  state = "OR",
  year = 2020,
  sumfile = "dhc"
)

# Compare with the ACS data
```
:::

# Using `blsAPI`

## Overview

The situation for BLS wrappers in R is a little more ambiguous, partly because the BLS API is less well-maintained than Census (the most recent update listed on their [website](https://www.bls.gov/bls/api_features.htm) is 2020). While `tidycensus` has become widely used, there are at least three different packages for accessing BLS data with various pros and cons.

1.  [`blsAPI`](https://github.com/mikeasilva/blsAPI) --- *Pros:* straightforward implementation, used in official BLS API documentation. *Cons:* mostly requires knowing specific series IDs for the data you want, as well as some knowledge of how to parse the returned data; most recent version is not on official CRAN repository.
2. [`blscrapeR`](https://github.com/keberwein/blscrapeR) --- *Pros:* a bit more user-friendly, compatible with the `tidyverse` style/format. *Cons:* unmaintained until just recently,  mostly requires knowing specific series IDs for the data you want.
3. [`BLSloadR`](https://schmidtdetr.github.io/BLSloadR/index.html) --- *Pros:* avoids the rate limits set by the BLS API, includes more intuitive functions, and is being developed by LMI shops. *Cons:* largely downloads data in bulk, is still in the early stages of development.

Recognizing these limitations, we at least wanted to examine the `blsAPI` package in part because it includes a more ergonomic function for retrieving **Local Area Unemployment Statistics (LAUS)** data, `laus_get_data()`. It also contains a similar function for retrieving QCEW data.

While the core `blsAPI()` function requires that you know the series ID of the data you want to retrieve, `laus_get_data` doesn't; you just provide location names, a measure, and a year range for the data you want.

Visit the package repository for full documentation: <https://github.com/mikeasilva/blsAPI>

### BLS API Versions and Limits

Before jumping in, it's also worth noting that the BLS API has two versions with different capabilities and different limits on how many API queries you make at one time, and how many data series you can return with a single query.

| Feature                    | Version 2.0 (Registered)  | Version 1.0 (Unregistered) |
|----------------------------|---------------------------|----------------------------|
| Daily query limit          | 500                       | 25                         |
| Series per query           | 50                        | 25                         |
| Years per query            | 20                        | 10                         |
| Net/Percent changes        | Yes                       | No                         |
| Annual averages            | Yes                       | No                         |

We will use **Version 2.0**, which requires a registration for a free API key. You can register for one at <https://data.bls.gov/registrationEngine/>.

## Getting Started

Because a compatible version of the package is only available via GitHub, we'll have to take a slightly different approach to installing it. 

```{r}
#| label: install-blsapi
#| eval: false

# First ensure that devtools package is installed:
install.packages("devtools")

# Now use the install_github function from devtools to install the blsAPI package.
# The package::function() syntax allows you to use a function without loading the entire package.
devtools::install_github("mikeasilva/blsAPI")
```

### Setting Up Your BLS API Key

Just like the Census API, you should store your BLS API key in your `.Renviron` file rather than hardcoding it in scripts:

```{r}
#| label: bls-key-setup
#| eval: false

# Open your .Renviron file for editing
# The usethis package should be installed automatically when you install devtools
usethis::edit_r_environ()

# Add this line to the file, then save:
# BLS_KEY=your_key_here

# Reload R environment 
readRenviron("~/.Renviron")

# Check that it is installed
Sys.getenv("BLS_KEY")
```

You can then retrieve your key in any script with `Sys.getenv("BLS_KEY")`. We will pass this key to `blsAPI` functions when making Version 2.0 requests.

```{r}
#| label: load-blsapi

library(blsAPI)
```

## Retrieving LAUS Data for Oregon Counties

The **Local Area Unemployment Statistics (LAUS)** program produces monthly employment and unemployment estimates for states, counties, metros, and other geographies. Let's use `laus_get_data()` to retrieve county-level data for Oregon.

### Step 1: Identify Oregon County Names

The `laus_get_data()` function takes location names as its first argument. These names need to match what the BLS uses in its area codes dataset. For counties, the format is `"County Name County, OR"`. You can explore the available area names using a dataset built into the `blsAPI` package:

```{r}
#| label: explore-area-codes
#| eval: false

# View a list of all available LAUS area names and their associated codes
View(LAUS_AreaCodes)

```

Let's define the Oregon counties we will work with. Oregon has 36 counties --- we will use a selection here to keep API requests manageable, since each call counts against the daily query limit:

```{r}
#| label: define-counties
#| eval: false

# Filter the LAUS area name to just those that contain "County, OR" and store as a vector object
or_counties <- LAUS_AreaCodes |>
  select(area_text) |>
  filter(str_detect(area_text, "County, OR"))

or_counties <- or_counties$area_text
```

::: {.callout-important}
## API Rate Limits
Each `laus_get_data()` call sends requests to the BLS API, which has daily query limits. With Version 2.0 (registered), you get 500 queries per day and up to 50 series per query. For a large number of counties, the function may need to make multiple API calls behind the scenes. If you are working with many counties and multiple measures, be mindful of your daily limit.
:::

### Step 2: Download Unemployment Rate and Labor Force Data

The `laus_get_data()` function takes four arguments: a vector of location names, the measure you want, a start year, and an end year. You need to make a **separate call for each measure**, since the function returns data for one measure at a time.

```{r}
#| label: laus-ur
#| eval: false

# Retrieve unemployment rate data for 2019-2024
or_ur <- laus_get_data(or_counties, "unemployment rate", 2019, 2024)

head(or_ur)
```

Notice the structure: the function returns a data frame with columns for `year`, `period`, `periodName`, the measure value (named based on your request --- here, `Unemployment_Rate`), and `Location`.

Now let's pull labor force data:

```{r}
#| label: laus-lf
#| eval: false

# Retrieve labor force data for the same counties and years
or_lf <- laus_get_data(or_counties, "labor force", 2019, 2024)

head(or_lf)
```

### Step 3: Calculate Average Annual Values

Let's create a summary table with average annual unemployment rate and labor force for each county. First, we need to clean up and reshape the data:

```{r}
#| label: laus-annual
#| eval: false

# Clean unemployment rate data: ensure numeric types and create a date column
or_ur_clean <- or_ur |>
  mutate(
    ur = as.numeric(Unemployment_Rate),
    year = as.numeric(year),
    date = as.Date(paste0(year, "-",
      str_pad(str_remove(period, "M"), 2, pad = "0"), "-01"))
  )

# Calculate average annual unemployment rate by county
or_ur_annual <- or_ur_clean |>
  filter(period != "M13") |>
  group_by(year, Location) |>
  summarize(avg_ur = round(mean(ur, na.rm = TRUE), 1), .groups = "drop")

# Clean and summarize labor force data the same way
or_lf_clean <- or_lf |>
  mutate(
    lf = as.numeric(Labor_Force),
    year = as.numeric(year)
  )

or_lf_annual <- or_lf_clean |>
  filter(period != "M13") |>
  group_by(year, Location) |>
  summarize(avg_lf = round(mean(lf, na.rm = TRUE), 0), .groups = "drop")

# Combine into one summary table
or_laus_annual <- or_ur_annual |>
  left_join(or_lf_annual, by = c("year", "Location")) |>
  mutate(county = str_remove(Location, ", OR$"))

head(or_laus_annual, 10)
```

::: {.callout-tip collapse="true"}
## Exercise 3: Explore the LAUS Data

Look at the `or_laus_annual` data frame. Which county in the dataset had the highest average unemployment rate in 2024? Which had the lowest? Use `filter()` and `arrange()` to find out.

```{r}
#| label: exercise-3
#| eval: false

# Highest unemployment rate in 2024
or_laus_annual |>
  filter(year == 2024) |>
  arrange(desc(avg_ur))

# Lowest unemployment rate in 2024
or_laus_annual |>
  filter(year == 2024) |>
  arrange(______)
```
:::

### Step 4: Create a Line Chart for One County

Let's visualize the monthly unemployment rate trend for a single county using the cleaned monthly data:

```{r}
#| label: fig-ur-single
#| fig-cap: "Monthly Unemployment Rate for Multnomah County, OR (2019–2024)"
#| fig-alt: "Line chart showing the unemployment rate trend for Multnomah County, Oregon from 2019 through 2024, with a visible spike during 2020."
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| eval: false

or_ur_clean |>
  filter(
    Location == "Multnomah County, OR",
    period != "M13"
  ) |>
  ggplot(aes(x = date, y = ur)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  labs(
    x = NULL,
    y = "Unemployment Rate",
    caption = "Source: Bureau of Labor Statistics, Local Area Unemployment Statistics"
  ) +
  theme_minimal(base_size = 12)
```

### Step 5: Faceted Line Chart for Multiple Counties

To compare several counties at once, we can use `facet_wrap()` in ggplot2 to create a small multiples chart:

```{r}
#| label: fig-ur-faceted
#| fig-cap: "Monthly Unemployment Rate for Selected Oregon Counties (2019–2024)"
#| fig-alt: "Faceted line chart showing unemployment rate trends for six Oregon counties from 2019 through 2024."
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
#| eval: false

# Select 6 counties for comparison
selected_counties <- c(
  "Multnomah County, OR",
  "Lane County, OR",
  "Deschutes County, OR",
  "Jackson County, OR",
  "Marion County, OR",
  "Coos County, OR"
)

or_ur_clean |>
  filter(
    Location %in% selected_counties,
    period != "M13"
  ) |>
  mutate(county = str_remove(Location, ", OR$")) |>
  ggplot(aes(x = date, y = ur)) +
  geom_line(color = "steelblue", linewidth = 0.7) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  facet_wrap(~ county, ncol = 3) +
  labs(
    x = NULL,
    y = "Unemployment Rate",
    caption = "Source: Bureau of Labor Statistics, Local Area Unemployment Statistics"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1, "lines")
  )
```

The faceted chart makes it easy to compare trends across counties while giving each one its own panel for clarity. Notice how all six counties show the dramatic spike during 2020 but with different magnitudes and recovery timelines.

::: {.callout-tip collapse="true"}
## Exercise 4: Your Own Faceted Chart

Create a faceted line chart showing the **labor force** (not unemployment rate) for a different set of 4--6 Oregon counties of your choice. You will need to work with the `or_lf` data frame instead of `or_ur_clean`.

```{r}
#| label: exercise-4
#| eval: false

# Step 1: Clean the labor force data (similar to what we did for unemployment rate)
my_counties <- c(
  # Pick 4-6 counties from the or_counties vector
)

or_lf_monthly <- or_lf |>
  mutate(
    lf = as.numeric(______),
    year = as.numeric(year),
    date = as.Date(paste0(year, "-",
      str_pad(str_remove(period, "M"), 2, pad = "0"), "-01"))
  )

# Step 2: Filter and plot
or_lf_monthly |>
  filter(
    Location %in% my_counties,
    period != "M13"
  ) |>
  mutate(county = str_remove(Location, ", OR$")) |>
  ggplot(aes(x = date, y = lf)) +
  geom_line(color = "steelblue", linewidth = 0.7) +
  facet_wrap(~ county, ncol = 2, scales = "free_y") +
  labs(
    x = NULL,
    y = "Labor Force",
    caption = "Source: Bureau of Labor Statistics, Local Area Unemployment Statistics"
  ) +
  theme_minimal()
```

**Bonus:** Notice the `scales = "free_y"` argument in `facet_wrap()`. What does it do? Try removing it and see how the chart changes. When would you want fixed vs. free scales?
:::

::: {.callout-tip collapse="true"}
## Exercise 5: Combine Census and BLS Data

This is a more advanced exercise that brings together both packages. Using the county-level population data you retrieved from tidycensus and the LAUS labor force data from blsAPI, calculate a **labor force participation rate proxy** for each Oregon county (average annual labor force divided by total population). Create a bar chart showing the results.

Hint: You will need to join the datasets on county name. Think about which year of Census data to pair with which year of LAUS data, and how to make the county name formats match between the two sources.

```{r}
#| label: exercise-5
#| eval: false

# Your code here
```
:::

### Using the Core blsAPI() Function

While `laus_get_data()` is convenient for LAUS data, the core `blsAPI()` function gives you access to *any* BLS series if you know the series ID. This is useful for programs that do not have a dedicated wrapper function.

Here is a quick example of requesting data with a known series ID:

```{r}
#| label: blsapi-direct
#| eval: false

# Request CPI-U (Consumer Price Index) data using API Version 2.0
payload <- list(
  "seriesid" = c("CUUR0000SA0"),
  "startyear" = 2020,
  "endyear" = 2024,
  "registrationkey" = Sys.getenv("BLS_KEY")
)

response <- blsAPI(payload, api_version = 2, return_data_frame = TRUE)
head(response)
```

You can find series IDs for any BLS program using the tools at <https://www.bls.gov/data/>.

## Other Available BLS Datasets

Beyond LAUS, the BLS produces many other datasets you can access through the API:

-   **CES** (Current Employment Statistics) --- employer-based estimates of employment, wages, and hours at state and metro levels
-   **OEWS** (Occupational Employment and Wage Statistics) --- employment and wage estimates by occupation
-   **JOLTS** (Job Openings and Labor Turnover Survey) --- data on job openings, hires, and separations
-   **CPS** (Current Population Survey) --- national household survey data with demographic detail
-   **QCEW** (Quarterly Census of Employment and Wages) --- comprehensive employer-reported data by industry and area. The blsAPI package includes a dedicated `blsQCEW()` function for this dataset.

Each of these can be accessed through `blsAPI()` using the appropriate series IDs, or through specialized wrapper functions where they exist.

# Wrapping Up

In this session we covered how to use R packages to access two major federal data sources directly from your scripts. The key takeaways are:

-   **APIs** let you retrieve data programmatically, making your workflow reproducible and scriptable. **Wrapper packages** like `tidycensus` and `blsAPI` hide the complexity of raw API calls and return ready-to-use data.
-   **tidycensus** gives you access to ACS, Decennial Census, Population Estimates, and more, with built-in support for spatial geometries.
-   **blsAPI** gives you access to LAUS and other BLS programs through the BLS API. The `laus_get_data()` convenience function lets you request LAUS data by location name and measure without constructing series IDs manually.

Both packages return data that integrates directly with the tidyverse and ggplot2 workflows you already know.